### 2010-01-04 article.tex
Building up solutions

\documentclass{article}

\usepackage{amsmath}
\usepackage{url}

\title{Optimizing Small Keyboard Layout to Minimize ``T9onyms''}
\author{P. Boothe}

\newcommand{\newword}[1]{{\bf #1}}
\newcommand{\NP}{\ensuremath{\mathcal{NP}}}
\newtheorem{theorem}{Theorem}

\begin{document}
\maketitle

\begin{abstract}
Sending text messages on cell phones which only contain the keys 0-9 and \#
and * can be a painful experience.  In order to make it easier, phone
manufactureres have designed predictive typing schemes.  In this paper we
consider multiple algorithms to design an optimal map of cell phone keys to
letters so that, for a given dictionary of words with associated frequencies,
the expected number of characters typed is minimized.  We end by describing the
best mapping we found, in the (vain) hope that it can become a new alternative
layout for cell phone keyboards.  Also, it is quite fun to think about how one
might optimize this system.
\end{abstract}

Typing a message using a keyboard which has fewer keys than the alphabet has
letters can be a painful task.  In an effort to minimize this pain, cellular
telephone manufacturers have created the ``T9'' input method, which attempts to
guess which letter (of the 3 or 4 possible) is intended when a user hits a
single key.  Enhancements of this input method also provide ``speculative
lookahead'' to report, at any given time, what word it is that the user is
most likely trying to enter and provide a completion.  Unfortunately, because there are fewer keys than there are letters in the English alphabet, there are words which have different spellings but the same input sequence.  As an example, in the traditional mapping of keys to characters ($abc\to 2$, $def\to3$, etc), both ``me'' and ``of'' have the input sequence ''63''.  Two words with the same input seuquence forces the user to press a third button to cycle through the possibilities in order from most likely (``of'') to least likely (``me'').  Even worse: ``home'', ``good'', ``gone'', ``hood'', ``hoof'', ``hone'', and ``goof'' all have the input sequence ``4663''.  If the * key is used as the ``next match'' button, then the user will actually have to type in ``4663******'' to type in the word ``goof''.

When two words have the same input sequence (neglecting the *'s at the end) then these words are \newword{t9onyms}.  When two or more words are t9onyms, then the less-popular words require extra keypresses, raising the expected number of keypresses to type in a random selection of words from the english dictionary.  In order to type a given word, one must press one button for every character in the word, followed by pressing the * key as many times as there are t9onyms which are more likely than the desired wored.  Because typing on a cellphone keyboard is already an unpleasant experience, we would like to minimize the expected number of keystrokes.  The number of keystrokes a word requires is equal to the number of letters in the word, plus the number of t9onyms which are more frequent than the word.  Formally, we define the {\sc MinimalKeystrokes} problem as:

\begin{theorem}[{\sc MinimumKeystrokes}]~\\
\label{thm:minstrokes}
{\sc Instance} An alphabet $A$, a set of words and their associated probabilities $W$, and a set of keys $K$ ($|A| > |K|$).\\
~\\
{\sc Question} What is the mapping of $A \to K$ which minimizes the expected number of characters to type a word from $W$?
\end{theorem}

and the corresponding decision problem is

\begin{theorem}~\\
{\sc Instance} An alphabet $A$, a set of words and their associated probabilities $W$, a set of keys $K$ ($|A| > |K|$), and a number $f$.\\
~\\
{\sc Question} Is there a mapping of $A \to K$ in which the expected number of characters to type a word from $W$ is less than $f$?
\end{theorem}

Before proceeding, we note that this we note that this problem is in \NP, as
any assignment may be verified to have and expected number of characters of
less than $f$ in time proportional to the total length of all the words in $W$.

It is highly likely this problem is \NP-complete, due to a reduction from one of the optimal trie problems.  But I'm not there yet.

However, our arbitrary remappings of the keyboard may be quickly deemed unusable.  Therefore, we define a refinement of the problem, in which the characters are required to be kept in alphabetical order as a sequence, and our only choices are about how to divide the subsequence into key assignments.  The default layout can be described as the partition $abc|def|ghi|jkl|mno|pqrs|tuv|wxyz$.  Is this partition optimal?  In order to decide this, we define the problem {\sc MinimumKeystrokePartition} as:
\begin{theorem}[{\sc MinimumKeystrokePartition}]~\\
\label{thm:minpartition}
{\sc Instance} A sequence of letters $A = \{a_1, a_2, \ldots \}$, a set of words and their associated probabilities $W$, and a set of keys $K = \{2, 3, 4, \ldots \}$ ($|A| > |K|$).\\
~\\
{\sc Question} What is the mapping $f$ of $A \to K$ which minimizes the expected number of characters to type a word from $W$, and where $f(a_1) = 2$ and if $f(a_i) = k$, then either $f(a_{i+1}) = k$ or $f(a_{i+1}) = k+1$?
\end{theorem}

As before, the corresponding decision problem is
\begin{theorem}~\\
{\sc Instance} A sequence of letters $A = \{a_1, a_2, \ldots \}$, a set of words and their associated probabilities $W$, a set of keys $K = \{2, 3, 4, \ldots \}$ ($|A| > |K|$), and a parameter $e$.\\
~\\
{\sc Question} Is there a mapping $f$ of $A \to K$ in which the expected number of characters to type a word from $W$ is less than $e$, and where $f(a_1) = 2$ and if $f(a_i) = k$, then $f(a_{i+1}) \in \{k, k+1\}$?
\end{theorem}

Again, we note that this decision problem is in \NP, as any proposed partition may be verified, in polynomial time, to have an expected value less than $e$.

\section{Exhaustive Search}

Because our problems are \NP-hard, we move on to combinatorial enumeration and experimental results.  The number of partitions of a set of size $n$, which we denote $p(n)$ is actually relatively small.  In particular, $p(26) = 2,436$\footnote{\url{http://www.wolframalpha.com/input/?i=PartitionsP[26]}}, which is easily enumerated and tested.
\end{document}

### 2010-01-04 alphabetpartitions.py

@@ -1,0 +1,22 @@
+#!/usr/bin/env python3
+def partitions(s, n):
+    "Generate all partitions of the given sequence"
+    if n == 0:
+        yield [ s ]
+    elif len(s) == 0:
+        yield []
+    elif len(s) == 1:
+        yield [ s ]
+    elif len(s) > 1:
+        yield [ s ]
+        for i in range(1,len(s)):
+            for p in partitions(s[i:], n-1):
+                yield [ s[:i] ] + p
+
+def alphabetpartition(n=8):
+    return partitions("abcdefghijklmnopqrstuvwxyz", 8)
+
+if __name__ == '__main__':
+    import sys
+    for p in alphabetpartition():
+        print(p)

### 2010-01-04 notr.cpp

@@ -1,0 +1,117 @@
+#include <algorithm>
+#include <iostream>
+#include <fstream>
+#include <map>
+#include <string>
+#include <vector>
+
+using namespace std;
+
+typedef pair<int, string> pis;
+vector<pis> words;
+long long total;
+
+bool piscompare(pis a, pis b) { return a.first > b.first; }
+
+long double quality(vector<string> &partition)
+{
+    map<char,string> rmap;
+    int i = 0;
+    string buttons = "23456789";
+    for (vector<string>::iterator s = partition.begin();
+            s != partition.end();
+            ++s, ++i) {
+        for (string::iterator c = s->begin(); c != s->end(); ++c) {
+            rmap[*c] = buttons[i];
+        }
+    }
+
+    map<string, vector<pis> > typing;
+    for (vector<pis>::iterator fw = words.begin();
+            fw != words.end();
+            ++fw) {
+        int freq = fw->first;
+        string word = fw->second;
+        string presses = "";
+        for (string::iterator c = word.begin(); c != word.end(); ++c) {
+            presses += rmap[*c];
+        }
+
+        if (typing.count(presses) == 0)
+            typing[presses] = vector<pis>();
+        typing[presses].push_back(*fw);
+    }
+
+    for (map<string, vector<pis> >::iterator w = typing.begin();
+            w != typing.end();
+            ++w) {
+        sort(w->second.begin(), w->second.end(), piscompare);
+    }
+
+    long long totalpresses = 0;
+    for (map<string, vector<pis> >::iterator w = typing.begin();
+            w != typing.end();
+            ++w) {
+        int wordsize = w->first.size();
+        for (int i = 0; i < w->second.size(); i++) {
+            totalpresses += (wordsize + i) * w->second[i].first;
+        }
+    }
+    return ((long double)totalpresses)/((long double) total);
+}
+
+int start = 0;
+int end = -1;
+void genallpartitions(string &s, int count, int index, 
+        vector<string> &part)
+{
+    if (count == 0) {
+        if (--start > 0) {
+            --end;
+        } else {
+            // Analyze it
+            cout << quality(part) << " ";
+            for (vector<string>::iterator i = part.begin();
+                    i != part.end();
+                    ++i)
+                cout << *i << " ";
+            cout << endl;
+            if (--end == 0) exit(0);
+        }
+    } else if (count == 1) {
[…]

### 2010-01-04 pcount.py

@@ -1,0 +1,16 @@
+#!/usr/bin/env python3
+def partitions(s):
+    "Generate all partitions of the given sequence"
+    if len(s) == 0:
+        yield []
+    elif len(s) == 1:
+        yield [ s ]
+    elif len(s) > 1:
+        yield [ s ]
+        for i in range(1,len(s)):
+            for p in partitions(s[i:]):
+                yield [ s[:i] ] + p
+
+import sys
+for p in partitions(list(range(int(sys.argv[1])))):
+    print(p)

### 2010-01-04 words/processall.py

@@ -1,0 +1,18 @@
+import sys
+count,word,_,_ = input().split()
+print(count,word)
+
+oldcount = count
+oldword = word
+for line in sys.stdin:
+    count,word,_,_ = line.split()
+    count = int(count)
+
+    if oldword == word:
+        oldcount += count
+    else:
+        if oldword.isalpha():
+            print(oldcount,oldword)
+        oldcount = count
+        oldword = word
+print(oldcount,oldword)

### 2010-01-12 a2.tex
Lots of changes.  Started in on the rewrite.

\documentclass{article}

\usepackage{amsmath}
\usepackage{url}

\newcommand{\newword}[1]{{\bf #1}}
\newtheorem{problem}{Problem}
\newcommand{\NP}{\ensuremath{\mathcal{NP}}}

\title{Using Cell Phone Keyboards is (\NP) Hard}
\author{P. Boothe}

\begin{document}
\maketitle

\begin{abstract}
Sending text messages on cell phones which only contain the keys 0-9 and
\# and * can be a painful experience.  We consider the problem of
designing an optimal mapping of numbers to sets of letters to act as an
alternative to the standard $\{2\to\{abc\}, 3\to\{def\}\ldots\}$.  Our
overall goal is to minimize the number of keys that must be typed.  We
prove that the problem is \NP hard, but describe several mappings which
improve the standard one.
\end{abstract}

Typing on a keyboard which has fewer keys than there are letter in
the alphabet can be a painful task.  There are a plethora of 
input schemes which attempt to make this task easier, but the one thing
they all have in common is that all of these input methods use the
standard mapping of numeric keys to alphabetic numbers of $\{2\to\{abc\},
         3\to\{def\}, 4\to\{ghi\}, 5\to\{jkl\}, 6\to\{mno\}, 7\to\{pqrs\},
         8\to\{tuv\}, 9\to\{wxyz\}\}$.  They use this mapping for historic
reasons, but, if we don't mind breaking some ``legacy'' applications of the
phone system such as ``1-800-FLOWERS'' and the like, then we might try
rearranging the numbers on the keys to make messages easier to type.

Before we get any farther, let us sketch the basic problem that we will keep
revising and revisiting throughout this paper.
\begin{problem}[{\sc MinimumKeystrokes}]~\\
{\sc Instance}: A set of letters corresponding to an alphabet $A$ $(|A| =
n)$, a number of keys $k$, an input method $IN$, and a set of
tuples of words and frequencies $W$.  The frequencies in $W$ are integers,
and the words are made up of solely of elements of $A$.  We will
treat $IN$ as a function which, given a partition of $A$ and a word
$w$, returns how many keystrokes are required to type $w$.~\\
        ~\\
{\sc Question}: What is the best partition of $A$ into $k$ sets, such that the
total number of keystrokes to type every word in $W$ its associated frequency
times is minimized?  Equivalently, what is the partition of $P$ of $A$ that
will minimize
$$\sum_{(w,f)\in W} f*IN(w,P)$$
\end{problem}

Over the course of this paper we consider three different real-world
schemes for $IN$ (basic typing, T9, and predictive T9), and for each
variant that is proven \NP-hard, we consider the restriction on $P$ that
requires that we keep the alphabet in alphabetical order.  In all cases
where it is computationally feasible, we provide results for the case of
the English language, on 8-key keyboards, using the British National
Corpus\cite{bnc}.

\bibliographystyle{abbrv}
\bibliography{bibliography}
\end{document}

### 2010-01-13 basicnopause.cpp
Better and better...

@@ -1,0 +1,59 @@
+#include <algorithm>
+#include <iostream>
+#include <fstream>
+#include <map>
+#include <string>
+#include <vector>
+
+using namespace std;
+
+
+int main(int argc, char** argv)
+{
+    int tmp;
+    ifstream win("words/wordlist");
+    win >> tmp;
+
+    string word;
+    win >> word; // Eat the rest of the header
+
+    map<char,int> presses;
+    presses['a'] = 1;
+    presses['b'] = 2;
+    presses['c'] = 3;
+    presses['d'] = 1;
+    presses['e'] = 2;
+    presses['f'] = 3;
+    presses['g'] = 1;
+    presses['h'] = 2;
+    presses['i'] = 3;
+    presses['j'] = 1;
+    presses['k'] = 2;
+    presses['l'] = 3;
+    presses['m'] = 1;
+    presses['n'] = 2;
+    presses['o'] = 3;
+    presses['p'] = 4;
+    presses['q'] = 3;
+    presses['r'] = 2;
+    presses['s'] = 1;
+    presses['t'] = 1;
+    presses['u'] = 2;
+    presses['v'] = 3;
+    presses['w'] = 1;
+    presses['x'] = 2;
+    presses['y'] = 3;
+    presses['z'] = 4;
+
+    long long total = 0;
+    int freq;
+    while (win >> freq) {
+        win >> word;
+        for (string::iterator i = word.begin(); i != word.end(); ++i)
+            total += freq * presses[*i];
+    }
+
+    cout << total << endl;
+
+    return 0;
+}

### 2010-01-13 a2.tex
Better and better...

@@
 \usepackage{url}
+\usepackage{graphicx}
 
@@
 \begin{abstract}
-Sending text messages on cell phones which only contain the keys 0-9 and
-\# and * can be a painful experience.  We consider the problem of
-designing an optimal mapping of numbers to sets of letters to act as an
-alternative to the standard $\{2\to\{abc\}, 3\to\{def\}\ldots\}$.  Our
-overall goal is to minimize the number of keys that must be typed.  We
-prove that the problem is \NP hard, but describe several mappings which
-improve the standard one.
+Sending text messages on cell phones which only contain the keys 0-9 and \# and
+* can be a painful experience.  We consider the problem of designing an optimal
+mapping of numbers to sets of letters to act as an alternative to the standard
+$\{2\to\{abc\}, 3\to\{def\}\ldots\}$.  Our overall goal is to minimize the
+number of buttons that must be pressed to enter an aferage message in English.
+We prove that the problem in almost all of its variations is \NP hard, but
+describe several mappings which improve the standard one.  With luck, one of
+these new mappings will achieve success similar to that of the Dvorak layout
+for computer keyboards.
 \end{abstract}
 
-Typing on a keyboard which has fewer keys than there are letter in
-the alphabet can be a painful task.  There are a plethora of 
-input schemes which attempt to make this task easier, but the one thing
-they all have in common is that all of these input methods use the
-standard mapping of numeric keys to alphabetic numbers of $\{2\to\{abc\},
+Typing on a keyboard which has fewer keys than there are letters in the
+alphabet can be a painful task.  There are a plethora of input schemes which
+attempt to make this task easier, but the one thing they all have in common is
+that all of these input methods use the standard mapping of numeric keys to
+alphabetic numbers of 
+$\{2\to\{abc\},
          3\to\{def\}, 4\to\{ghi\}, 5\to\{jkl\}, 6\to\{mno\}, 7\to\{pqrs\},
-         8\to\{tuv\}, 9\to\{wxyz\}\}$.  They use this mapping for historic
-reasons, but, if we don't mind breaking some ``legacy'' applications of the
-phone system such as ``1-800-FLOWERS'' and the like, then we might try
-rearranging the numbers on the keys to make messages easier to type.
+         8\to\{tuv\}, 9\to\{wxyz\}\}$.
+However, if we don't mind breaking some ``legacy'' applications of the phone
+system such as ``1-800-FLOWERS'', then we might try rearranging the numbers on
+the keys to make messages easier to type.  Most variants of this problem turn
+out to be NP-hard, unfortunately.
 
@@
 total number of keystrokes to type every word in $W$ its associated frequency
-times is minimized?  Equivalently, what is the partition of $P$ of $A$ that
+times is minimized?  Equivalently, what is the partition of $P$ of $A$ ($|P| = k$) that
 will minimize
 $$\sum_{(w,f)\in W} f*IN(w,P)$$
+\label{probtemplate}
 \end{problem}
@@
 requires that we keep the alphabet in alphabetical order.  In all cases
-where it is computationally feasible, we provide results for the case of
+where it is computationally feasible we provide results for the case of
 the English language, on 8-key keyboards, using the British National
 Corpus\cite{bnc}.
+
+\section{Setting a Baseline and Solving the Easiest Problem}
+
+The baseline we compare against is the most painful method of text entry.  I
+imagine that no ``digital native'' actually uses this input method, and that it
+is available on all phones solely that their parents might use it.  In this
+method, to type an `a', the user of the cellphone types the `2' key; to type a
+`b' they type `22', to type `c' they type `222', to type a `d' they type `3',
+to type an `e' they type `33', and so on.  To type a word with multiple
+letters that use the same key, such as `accept', the user must pause
+between keystrokes.  Thus, the full key entry sequence for `accept', using
+`.' to indicate a pause, is `2.22.223378'.
+
+\begin{figure}
+\includegraphics[width=2in]{phonekeys.jpg}
+\caption{A cell phone keyboard}
+\label{keypic}
+\end{figure}
+
+Initially, we will completely neglect the pauses and concern ourselves solely
+with the number of keystrokes required.  This implies that `a' will always
+require one button press, `b' will always require two, and so on.  Thus, to get
+a baseline of how many keystrokes are required to enter the entirety of our
+corpus of words $W$, we count the total number of occurences of each letter,
+and multiply that number of occurrences by the number of button pushes
+required by that letter.  Running this on the British National Corpus
+using the standard cell phone keyboard layout, we find that entering the
+entirety of the 100,106,029 word occurrences in the corpus would require
+948,049,046 button presses.
+
+If we examine a cell phone keyboard (Figure~\ref{keypic}) then we can see that
+there are some terrible choices being made.  The frequently-occurring letters
+'r' and 's' require more keystrokes than 'q'!  Surely, a better designed
+keyboard can do better than this.  If we just reversed the order of the letters
+on the 7 key, from `pqrs' to `srqp', the number of button presses required
+would drop to 865,118,331 --- a savings of more than $8.7\%$!  If we assume
+that every button press takes an equal amount of time, then this corresponds
+the users spending $8.7\%$ less time entering their text messages.  In an
+effort to find out how low this can go, we define the problem {\sc
+BasicCellPhoneTyping} based on the template problem on
+page~\pageref{probtemplate}.
 

### 2010-01-15 a2.tex
Easy version: knocked out of the park (for now).

@@
 
-\usepackage{amsmath}
+\usepackage{amsmath,amsthm}
 \usepackage{url}
@@
 \newtheorem{problem}{Problem}
+\newtheorem{theorem}{Theorem}
 \newcommand{\NP}{\ensuremath{\mathcal{NP}}}
@@
 the English language, on 8-key keyboards, using the British National
-Corpus\cite{bnc}.
+Corpus\cite{bnc} (BNC).  One feature of note is that no matter what scheme we
+use, the problem is trivial if the number of keys is not smaller than the
+number of letters in the alphabet ($k \ge |A|$).  Using tiny keyboards is only
+(computationally) difficult when the keyboards don't have enough keys.
 
-\section{Setting a Baseline and Solving the Easiest Problem}
+\section{Setting a Baseline with the Easiest Problem}
 
@@
 
+\begin{problem}[{\sc BasicCellPhoneTyping}]~\\
+{\sc Instance}: A set of letters corresponding to an alphabet $A$ $(|A| =
+n)$, a number of keys $k$, and a set of
+tuples of words and frequencies $W$.  The frequencies in $W$ are integers,
+and the words are made up of solely of elements of $A$. \\
+        ~\\
+{\sc Question}: What is the best partition of $A$ into $k$ sequences, such that the
+total number of button presses to type every word in $W$ its associated frequency
+times is minimized?  The number of button presses is equal to the order of the
+letter in the sequence assigned to a given key.  For example, in the key
+$2\to\{abc\}$, $a$ requires one keystroke and $c$ requires 3.
+\label{probtemplate}
+\end{problem}
+
+Unlike almost every other problem dealing with key assignment, we can construct
+an efficient algorithm for this problem.  We can construct a provable optimal
+greedy algorithm which works in time $O(|W| + |A| \log |A|)$.  Our algorithm is
+detailed in Figure~\ref{greedyalg}, and involves creating a histogram of the
+letters, and then assigning letters to keys round-robin style in the order from
+most-popular to least-popular.  To prove optimality, we invoke an exchange argument.
+
+\begin{theorem}
+The greedy algorithm finds an optimal solution to {\sc BasicCellPhoneTyping}.
+\label{basicthm}
+\end{theorem}
+\begin{proof}
+We will assume, for simplicity of proof, that all letters occur a different number of times in the corpus.  
+We begin by comparing two assignments of letters to keys by, for each assignment,
+constructing a sequence of sets, or spectrum, $S$.  The set $S_1$ (layer 1) is the set of all
+letters which require a single button press (that is, they are the first
+letter in their sequence on their assigned key), $S_2$ (layer 2) is the set of
+all letters which require two button presses, and so on.  If two assignments
+have the same sequence of sets, then one assignment may be transformed into
+the other assignment simply by swapping letter between keys without
+increasing or decreasing the total number of button pushes required to enter
+the corpus of words.  If two sequences have equal spectra, then we say their corresponding assignments are isomorphic.
+
+Now assume for the sake of contradiction that we have assignment $D$, with spectrum $T$, that is not isomorphic to the greedy solution $G$ with spectrum $S$.
+Note that in spectrum resulting from the greedy algorithm, all letters at layer
+$i$ occur strictly more frequently than the letters at layer $j$, $j > i$.
+Also, in $S$, for all layers $i$ except for the last layer, $|S_i| = k$.
+Because $T$ and $S$ are not isomorphic, in $T$ there must be at least one layer $i$ such that $T_i \neq S_i$.  Let $T_i$ be the first layer for which that is true.
+
+There are two cases for layer $T_i$.  In the first case, $T_i < k$, and we may
+remove any element from a later layer and create a strictly better layout.  In
+the second case, there is some letter $a$ such that $a \in T_i$ and $a \not\in
+S_i$.  We choose the least-frequent such $a$.  Because the greedy layout
+algorithm creates layers in order of frequency, we know that the frequency of
+$a$ is less than that of some letter $b$ in layer $T_j$, $j>i$.  Therefore, by
+creating a new layout with $a$ and $b$ swapped between $T_i$ and $T_j$, we have
+strictly decreased the number of button presses.  Therefore, in both cases,
+$T$ was not an optimal layout, and we have reached our contradiction.
+\end{proof}
+
+For the BNC, the optimal layout has the spectrum 
+$$[\{etaoinsr\}, \{hldcumfp\}, \{gwybvkxj\}, \{qz\}]$$
+and will require 638,276,114 button presses to
+entire the entire corpus instead of our original requirements of
+948,049,046.  This represents a savings of $32.67\%$ over our original
+layout, but only for the case of users who type using the most basic of
+input methods.  Unfortunately, this is the group of users who are likely
+the least adaptable to change, as almost all ``digital natives'' use one of
+the predictive methods to input text messages\footnote{I have no statistics
+on this except for an informal survey of one of my classes.  In that
+class, every student who used SMS either had a cellphone with a 26+
+key alphabetic keyboard or used a predictive method.}.  If we would
+like to place the least burden on these users whilst still
+decreasing the number of button presses, we now consider schemes
+where the only alteration of the keyboard is the rearrangement of
+the sequence of letters for a given key.
+
+By an argument symmetric to the proof of Theorem~\ref{basicthm}, we
+find that the optimal layout in this new scheme is to place the letters of a
+given key in sorted order.  Thus, the keyboard layout changes to $2\to[acb],
+3\to[edf], 4\to[ihg], 5\to[lkj], 6\to[onm], 7\to[srpq],
+8\to[tuv], 9\to[wyxz]$ and requires 678,547,463 button presses to
+enter the whole corpus.  This represents a $28.43\%$ savings, but has the
+added benefit that it does not change which key is mapped to which set of
+letters.  Furthermore, it also is \newword{legacy preserving}, as
+creating a keyboard with this layout will not invalidate such advertising
+gems as 1-800-FLOWERS or 1-800-MARINES.  Thus, we can speed up users by
+$28.43\%$ (neglecting inter-letter pauses) and not undermine 50 years of
+phone advertising.  This layout represents perhaps the most plausible
+layout yet.
+
 \bibliographystyle{abbrv}

### 2010-01-15 basic.cpp

@@ -1,0 +1,86 @@
+#include <algorithm>
+#include <iostream>
+#include <fstream>
+#include <map>
+#include <string>
+#include <vector>
+
+using namespace std;
+
+bool freqcompare(pair<char, long long> f, pair<char, long long> s) { return f.second > s.second; }
+
+int main(int argc, char** argv)
+{
+    int tmp;
+    ifstream win("words/wordlist");
+    win >> tmp;
+
+    string word;
+    win >> word; // Eat the rest of the header
+
+    map<char,long long> freq;
+    freq['a'] = 0;
+    freq['b'] = 0;
+    freq['c'] = 0;
+    freq['d'] = 0;
+    freq['e'] = 0;
+    freq['f'] = 0;
+    freq['g'] = 0;
+    freq['h'] = 0;
+    freq['i'] = 0;
+    freq['j'] = 0;
+    freq['k'] = 0;
+    freq['l'] = 0;
+    freq['m'] = 0;
+    freq['n'] = 0;
+    freq['o'] = 0;
+    freq['p'] = 0;
+    freq['q'] = 0;
+    freq['r'] = 0;
+    freq['s'] = 0;
+    freq['t'] = 0;
+    freq['u'] = 0;
+    freq['v'] = 0;
+    freq['w'] = 0;
+    freq['x'] = 0;
+    freq['y'] = 0;
+    freq['z'] = 0;
+
+    long long count;
+    while (win >> count) {
+        win >> word;
+        for (string::iterator i = word.begin(); i != word.end(); ++i)
+            freq[*i] += count;
+    }
+
+    vector<pair<char,long long> > ordered;
+    for (map<char,long long>::iterator i = freq.begin();
+            i != freq.end();
+            ++i)
+        ordered.push_back(*i);
+    sort(ordered.begin(), ordered.end(), freqcompare);
+
+    for (vector<pair<char,long long> >::iterator i = ordered.begin();
+            i != ordered.end();
+            i++)
+        cout << "(" << i->first << ", " << i->second << ") ";
+    cout << endl;
+
+    count = 0;
+    int level = 0;
+    int levelcount = 0;
+    for (vector<pair<char,long long> >::iterator i = ordered.begin();
+            i != ordered.end();
+            i++, levelcount++) {
+        if (levelcount % 8 == 0) {
+            cout << endl << "Level " << ++level << ": ";
+        }
+
+        cout << "(" << i->first << ", " << i->second << ") ";
[…]

### 2010-01-15 basicreorder.cpp
Easy version: knocked out of the park (for now).

@@ -1,0 +1,94 @@
+#include <algorithm>
+#include <iostream>
+#include <fstream>
+#include <map>
+#include <string>
+#include <vector>
+
+using namespace std;
+
+bool freqcompare(pair<char, long long> f, pair<char, long long> s) { return f.second > s.second; }
+
+int main(int argc, char** argv)
+{
+    int tmp;
+    ifstream win("words/wordlist");
+    win >> tmp;
+
+    string word;
+    win >> word; // Eat the rest of the header
+
+    map<char,long long> freq;
+    freq['a'] = 0;
+    freq['b'] = 0;
+    freq['c'] = 0;
+    freq['d'] = 0;
+    freq['e'] = 0;
+    freq['f'] = 0;
+    freq['g'] = 0;
+    freq['h'] = 0;
+    freq['i'] = 0;
+    freq['j'] = 0;
+    freq['k'] = 0;
+    freq['l'] = 0;
+    freq['m'] = 0;
+    freq['n'] = 0;
+    freq['o'] = 0;
+    freq['p'] = 0;
+    freq['q'] = 0;
+    freq['r'] = 0;
+    freq['s'] = 0;
+    freq['t'] = 0;
+    freq['u'] = 0;
+    freq['v'] = 0;
+    freq['w'] = 0;
+    freq['x'] = 0;
+    freq['y'] = 0;
+    freq['z'] = 0;
+
+    long long count;
+    while (win >> count) {
+        win >> word;
+        for (string::iterator i = word.begin(); i != word.end(); ++i)
+            freq[*i] += count;
+    }
+
+    vector<string> keys;
+    keys.push_back("abc");
+    keys.push_back("def");
+    keys.push_back("ghi");
+    keys.push_back("jkl");
+    keys.push_back("mno");
+    keys.push_back("pqrs");
+    keys.push_back("tuv");
+    keys.push_back("wxyz");
+
+
+    count = 0;
+    int key = 2;
+    for (vector<string>::iterator s = keys.begin();
+            s != keys.end();
+            ++s, ++key) {
+        cout << key << "\\to [";
+        vector<pair<char,long long> > ordered;
+        for (string::iterator c = s->begin();
+                c != s->end();
+                ++c)
+            ordered.push_back(pair<char,long long>(*c, freq[*c]));
+        sort(ordered.begin(), ordered.end(), freqcompare);
+
+        int j = 1;
+        for (vector<pair<char,long long> >::iterator i = ordered.begin();
+                i != ordered.end();
[…]

### 2010-01-15 basic.cpp
Little cleanups.

@@ -80,7 +80,7 @@ int main(int argc, char** argv)
         count += i->second * level;
     }
     cout << endl;
-    cout << count << endl;
+    cout << count << " button presses" << endl;
 
     return 0;
 }

### 2010-01-15 intdata/normalize.py
Little cleanups.

@@ -1,0 +1,6 @@
+import sys
+import math
+base = int(input())
+print(1)
+for line in sys.stdin:
+    print(int(line)/base)

### 2010-01-15 article.tex
Title changed
Got article.tex to be the actual article I am writing.  Hooray!

@@
 
-\usepackage{amsmath}
+\usepackage{amsmath,amsthm}
 \usepackage{url}
-
-\title{Optimizing Small Keyboard Layout to Minimize ``T9onyms''}
-\author{P. Boothe}
+\usepackage{graphicx}
 
 \newcommand{\newword}[1]{{\bf #1}}
+\newtheorem{problem}{Problem}
+\newtheorem{theorem}{Theorem}
 \newcommand{\NP}{\ensuremath{\mathcal{NP}}}
-\newtheorem{theorem}{Theorem}
+
+\title{Using Cell Phone Keyboards is (\NP) Hard}
+\author{P. Boothe}
 
@@
 \begin{abstract}
-Sending text messages on cell phones which only contain the keys 0-9 and \#
-and * can be a painful experience.  In order to make it easier, phone
-manufactureres have designed predictive typing schemes.  In this paper we
-consider multiple algorithms to design an optimal map of cell phone keys to
-letters so that, for a given dictionary of words with associated frequencies,
-the expected number of characters typed is minimized.  We end by describing the
-best mapping we found, in the (vain) hope that it can become a new alternative
-layout for cell phone keyboards.  Also, it is quite fun to think about how one
-might optimize this system.
+Sending text messages on cell phones which only contain the keys 0-9 and \# and
+* can be a painful experience.  We consider the problem of designing an optimal
+mapping of numbers to sets of letters to act as an alternative to the standard
+$\{2\to\{abc\}, 3\to\{def\}\ldots\}$.  Our overall goal is to minimize the
+number of buttons that must be pressed to enter an aferage message in English.
+We prove that the problem in almost all of its variations is \NP\ hard, but
+describe several mappings which improve the standard one.  With luck, one of
+these new mappings will achieve success similar to that of the Dvorak layout
+for computer keyboards.
 \end{abstract}
 
-Typing a message using a keyboard which has fewer keys than the alphabet has
-letters can be a painful task.  In an effort to minimize this pain, cellular
+Typing on a keyboard which has fewer keys than there are letters in the
+alphabet can be a painful task.  There are a plethora of input schemes which
+attempt to make this task easier, but the one thing they all have in common is
+that all of these input methods use the standard mapping of numeric keys to
+alphabetic numbers of 
+$\{2\to\{abc\},
+         3\to\{def\}, 4\to\{ghi\}, 5\to\{jkl\}, 6\to\{mno\}, 7\to\{pqrs\},
+         8\to\{tuv\}, 9\to\{wxyz\}\}$.
+In this paper, we consider schemes of rearranging the numbers on the keys to
+make messages easier to type.  Most variants of this problem turn out to be
+NP-hard, unfortunately.
+
+Before we get any farther, let us sketch the basic problem that we will keep
+revising and revisiting throughout this paper.
+\begin{problem}[{\sc MinimumKeystrokes}]~\\
+{\sc Instance}: A set of letters corresponding to an alphabet $A$ $(|A| =
+n)$, a number of keys $k$, an input method $IN$, and a set of
+tuples of words and frequencies $W$.  The frequencies in $W$ are integers,
+and the words are made up of solely of elements of $A$.  We will
+treat $IN$ as a function which, given a partition of $A$ and a word
+$w$, returns how many keystrokes are required to type $w$.~\\
+        ~\\
+{\sc Question}: What is the best partition of $A$ into $k$ sets, such that the
+total number of keystrokes to type every word in $W$ its associated frequency
+times is minimized?  Equivalently, what is the partition of $P$ of $A$ ($|P| = k$) that
+will minimize
+$$\sum_{(w,f)\in W} f*IN(w,P)$$
+\label{probtemplate}
+\end{problem}
+
+Over the course of this paper we consider three different real-world
+schemes for $IN$ (basic typing, T9, and predictive T9), and for each
+variant that is proven \NP-hard, we consider the restriction on $P$ that
+requires that we keep the alphabet in alphabetical order.  In all cases
+where it is computationally feasible we provide results for the case of
+the English language, on 8-key keyboards, using the British National
+Corpus\cite{bnc} (BNC).  One feature of note is that no matter what scheme we
+use, the problem is trivial if the number of keys is not smaller than the
+number of letters in the alphabet ($k \ge |A|$).  Using tiny keyboards is only
+(computationally) difficult when the keyboards don't have enough keys.
+
+\section{Setting a Baseline with the Easiest Problem}
+
+The baseline we compare against is the most painful method of text entry.  I
+imagine that no ``digital native'' actually uses this input method, and that it
+is available on all phones solely that their parents might use it.  In this
+method, to type an `a', the user of the cellphone types the `2' key; to type a
+`b' they type `22', to type `c' they type `222', to type a `d' they type `3',
+to type an `e' they type `33', and so on.  To type a word with multiple
+letters that use the same key, such as `accept', the user must pause
+between keystrokes.  Thus, the full key entry sequence for `accept', using
+`.' to indicate a pause, is `2.22.223378'.
+
+\begin{figure}
+\includegraphics[width=2in]{phonekeys.jpg}
+\caption{A cell phone keyboard}
+\label{keypic}
+\end{figure}
+
+Initially, we will completely neglect the pauses and concern ourselves solely
+with the number of keystrokes required.  This implies that `a' will always
+require one button press, `b' will always require two, and so on.  Thus, to get
+a baseline of how many keystrokes are required to enter the entirety of our
+corpus of words $W$, we count the total number of occurences of each letter,
+and multiply that number of occurrences by the number of button pushes
+required by that letter.  Running this on the British National Corpus
+using the standard cell phone keyboard layout, we find that entering the
+entirety of the 100,106,029 word occurrences in the corpus would require
+948,049,046 button presses.
+
+If we examine a cell phone keyboard (Figure~\ref{keypic}) then we can see that
+there are some terrible choices being made.  The frequently-occurring letters
+'r' and 's' require more keystrokes than 'q'!  Surely, a better designed
+keyboard can do better than this.  If we just reversed the order of the letters
+on the 7 key, from `pqrs' to `srqp', the number of button presses required
+would drop to 865,118,331 --- a savings of more than $8.7\%$!  If we assume
+that every button press takes an equal amount of time, then this corresponds
+the users spending $8.7\%$ less time entering their text messages.  In an
+effort to find out how low this can go, we define the problem {\sc
+BasicCellPhoneTyping} based on the template problem on
+page~\pageref{probtemplate}.
+
+\begin{problem}[{\sc BasicCellPhoneTyping}]~\\
+{\sc Instance}: A set of letters corresponding to an alphabet $A$ $(|A| =
+n)$, a number of keys $k$, and a set of
+tuples of words and frequencies $W$.  The frequencies in $W$ are integers,
+and the words are made up of solely of elements of $A$. \\
+        ~\\
+{\sc Question}: What is the best partition $P$ of $A$ into $k$ sequences, such
+that the total number of button presses to type every word in $W$ its
+associated frequency times is minimized?  The number of button presses is equal
+to the order of the letter in the sequence assigned to a given key.  For
+example, in the key $2\to\{abc\}$, $a$ requires one keystroke and $c$ requires
+3.  Equivalently, if we denote the number of button presses required to type
+letter $a$ with partition $P$ as $IN_P(a)$, then we want to find the $P$ that
+minimizes $$\sum_{(w,f) \in W}\sum_{c\in w} IN_P(c) * f$$
+\label{bcpt}
+\end{problem}
+
+Unlike almost every other problem dealing with key assignment, we can construct
+an efficient algorithm for this problem.  We can construct a provable optimal
+greedy algorithm which works in time $O(|W| + |A| \log |A|)$.  Our algorithm is
+detailed in Figure~\ref{greedyalg}, and involves creating a histogram of the
+letters, and then assigning letters to keys round-robin style in the order from
+most-popular to least-popular.  To prove optimality, we invoke an exchange argument.
+
+\begin{theorem}
+The greedy algorithm finds an optimal solution to {\sc BasicCellPhoneTyping}.
+\label{basicthm}
+\end{theorem}
+\begin{proof}
+We will assume, for simplicity of proof, that all letters occur a different number of times in the corpus.  
+We begin by comparing two assignments of letters to keys by, for each assignment,
+constructing a sequence of sets, or spectrum, $S$.  The set $S_1$ (layer 1) is the set of all
+letters which require a single button press (that is, they are the first
+letter in their sequence on their assigned key), $S_2$ (layer 2) is the set of
+all letters which require two button presses, and so on.  If two assignments
+have the same sequence of sets, then one assignment may be transformed into
+the other assignment simply by swapping letter between keys without
+increasing or decreasing the total number of button pushes required to enter
+the corpus of words.  If two sequences have equal spectra, then we say their corresponding assignments are isomorphic.
+
+Now assume for the sake of contradiction that we have assignment $D$, with spectrum $T$, that is not isomorphic to the greedy solution $G$ with spectrum $S$.
+Note that in spectrum resulting from the greedy algorithm, all letters at layer
+$i$ occur strictly more frequently than the letters at layer $j$, $j > i$.
+Also, in $S$, for all layers $i$ except for the last layer, $|S_i| = k$.
+Because $T$ and $S$ are not isomorphic, in $T$ there must be at least one layer $i$ such that $T_i \neq S_i$.  Let $T_i$ be the first layer for which that is true.
+
+There are two cases for layer $T_i$.  In the first case, $T_i < k$, and we may
+remove any element from a later layer and create a strictly better layout.  In
+the second case, there is some letter $a$ such that $a \in T_i$ and $a \not\in
+S_i$.  We choose the least-frequent such $a$.  Because the greedy layout
+algorithm creates layers in order of frequency, we know that the frequency of
+$a$ is less than that of some letter $b$ in layer $T_j$, $j>i$.  Therefore, by
+creating a new layout with $a$ and $b$ swapped between $T_i$ and $T_j$, we have
+strictly decreased the number of button presses.  Therefore, in both cases,
+$T$ was not an optimal layout, and we have reached our contradiction.
+\end{proof}
+
+For the BNC, the optimal layout has the spectrum 
+$$[\{etaoinsr\}, \{hldcumfp\}, \{gwybvkxj\}, \{qz\}]$$
+and any layout with that spectrum requires 638,276,114 button presses to
+entire the entire BNC instead of our original requirements of
+948,049,046.  This represents a savings of $32.67\%$ over our original
+layout, but only for the users who type using this most basic of
+input methods.  Unfortunately, this is the group of users who are likely
+the least adaptable to change, as almost all ``digital natives'' use 
+predictive methods to input text messages\footnote{I have no statistics
+on this except for an informal survey of one of my classes.  In that
+class, every student who used SMS either had a cellphone with a 26+
+key alphabetic keyboard or used a predictive method.}.  In order to place the
+least burden of ``newness'' on these users whilst still decreasing the number
+of button presses, we now consider schemes where the only alteration of the
+keyboard is the rearrangement of the sequence of letters for a given key.
+
+By an argument symmetric to the proof of Theorem~\ref{basicthm}, we find that
+the optimal layout in this new scheme is to place the letters of a given key in
+sorted order.  Thus, the keyboard layout changes to $2\to[acb], 3\to[edf],
+4\to[ihg], 5\to[lkj], 6\to[onm], 7\to[srpq], 8\to[tuv], 9\to[wyxz]$ and
+requires 678,547,463 button presses to enter the whole corpus.  This represents
+a $28.43\%$ savings, but has the added benefit that it does not change which
+key is mapped to which set of letters.  This makes it \newword{legacy
+preserving}, as creating a keyboard with this layout will not invalidate such
+advertising gems as 1-800-FLOWERS.  Thus, we can speed up users by $28.43\%$
+(neglecting inter-letter pauses) and not undermine 50 years of advertising.
+This layout represents perhaps the most plausible layout yet\footnote{The
+described layout is the only legacy preserving layout in this paper, and
+therefore should be considered the most practical suggestion, despite the fact
+that its algorithm is among the least FUN.  It also has the added benefit of
+not angering any of the organizations behind the numbers 1-800-FLOWERS,
+1-800-THRIFTY, and 1-800-MARINES}.  After setting up this baseline for the easy
+problem, we turn our attention to optimizing cellphone keyboards for predictive
+input methods.
+
+\section{The T9 Input Method}
+
+In an effort to minimize the pain of cellphone keyboard typing, cellular
 telephone manufacturers have created the ``T9'' input method, which attempts to
@@
 single key.  Enhancements of this input method also provide ``speculative
-lookahead'' to report, at any given time, what word it is that the user is
-most likely trying to enter and provide a completion.
-Unfortunately, because there are fewer keys than there are letters in the English alphabet, there are words which have different spellings but the same input sequence.
-As an example, in the traditional mapping of keys to characters ($abc\to 2$, $def\to3$, etc), both ``me'' and ``of'' have the input sequence ''63''.
-Two words with the same input seuquence forces the user to press a third button to cycle through the possibilities in order from most likely (``of'') to least likely (``me'').
-Even worse: ``home'', ``good'', ``gone'', ``hood'', ``hoof'', ``hone'', and ``goof'' all have the input sequence ``4663''.
-If the * key is used as the ``next match'' button, then the user will actually have to type in ``4663******'' to type in the word ``goof''.
+lookahead'' to report, at any given time, what word it is that the user is most
+likely trying to enter and provide a completion.  Unfortunately, because there
+are fewer keys than there are letters in the English alphabet, there are words
+which have different spellings but the same input sequence.  As an example, in
+the traditional mapping of keys to characters ($abc\to 2$, $def\to3$, etc),
+both ``me'' and ``of'' have the input sequence ''63''.  Two words with the same
+input sequence force the user to press a third button to cycle through the
+possibilities in order from most likely (``of'') to least likely (``me'').
+Even worse: ``home'', ``good'', ``gone'', ``hood'', ``hoof'', ``hone'', and
+``goof'' all have the input sequence ``4663''.  If the * key is used as the
+``next match'' button, then the user will actually have to type in
+``4663******'' to type in the word ``goof''.
 
@@
 In particular, $p(26) = 2,436$\footnote{\url{http://www.wolframalpha.com/input/?i=PartitionsP[26]}}, which is easily enumerated and tested.
+
+\bibliographystyle{abbrv}
+\bibliography{bibliography}
 \end{document}

### 2010-01-15 basicreorder.cpp

@@ -9,6 +9,8 @@ using namespace std;
 
 bool freqcompare(pair<char, long long> f, pair<char, long long> s) { return f.second > s.second; }
 
+string alphabet = "abcdefghijklmnopqrstuvwxyz";
+
 int main(int argc, char** argv)
 {
     int tmp;
@@ -19,32 +21,10 @@ int main(int argc, char** argv)
     win >> word; // Eat the rest of the header
 
     map<char,long long> freq;
-    freq['a'] = 0;
-    freq['b'] = 0;
-    freq['c'] = 0;
-    freq['d'] = 0;
-    freq['e'] = 0;
-    freq['f'] = 0;
-    freq['g'] = 0;
-    freq['h'] = 0;
-    freq['i'] = 0;
-    freq['j'] = 0;
-    freq['k'] = 0;
-    freq['l'] = 0;
-    freq['m'] = 0;
-    freq['n'] = 0;
-    freq['o'] = 0;
-    freq['p'] = 0;
-    freq['q'] = 0;
-    freq['r'] = 0;
-    freq['s'] = 0;
-    freq['t'] = 0;
-    freq['u'] = 0;
-    freq['v'] = 0;
-    freq['w'] = 0;
-    freq['x'] = 0;
-    freq['y'] = 0;
-    freq['z'] = 0;
+    for (string::iterator c = alphabet.begin();
+            c != alphabet.end();
+            ++c)
+        freq[*c] = 0;
 
     long long count;
     while (win >> count) {

### 2010-01-15 basicnopause.cpp
Got article.tex to be the actual article I am writing.  Hooray!

@@ -7,7 +7,6 @@
 
 using namespace std;
 
-
 int main(int argc, char** argv)
 {
     int tmp;
@@ -33,10 +32,10 @@ int main(int argc, char** argv)
     presses['m'] = 1;
     presses['n'] = 2;
     presses['o'] = 3;
-    presses['p'] = 4;
-    presses['q'] = 3;
-    presses['r'] = 2;
-    presses['s'] = 1;
+    presses['p'] = 1;
+    presses['q'] = 2;
+    presses['r'] = 3;
+    presses['s'] = 4;
     presses['t'] = 1;
     presses['u'] = 2;
     presses['v'] = 3;

### 2010-01-15 article.tex
Uses LLNCS format, and the makefile has a clean target.

@@
-\documentclass{article}
+\documentclass[runningheads]{llncs}
 
-\usepackage{amsmath,amsthm}
+\usepackage{amsmath}
 \usepackage{url}
 \usepackage{graphicx}
+\usepackage{salgorithm}
 
 \newcommand{\newword}[1]{{\bf #1}}
-\newtheorem{problem}{Problem}
-\newtheorem{theorem}{Theorem}
 \newcommand{\NP}{\ensuremath{\mathcal{NP}}}
+\spnewtheorem{prob}[theorem]{Problem}{\bfseries}{\itshape}
 
 \title{Using Cell Phone Keyboards is (\NP) Hard}
-\author{P. Boothe}
+\author{Peter Boothe}
+\institute{Manhattan College\\
+\email{peter.boothe@manhattan.edu}
+}
 
@@
 revising and revisiting throughout this paper.
-\begin{problem}[{\sc MinimumKeystrokes}]~\\
+\begin{prob}[{\sc MinimumKeystrokes}]~\\
 {\sc Instance}: A set of letters corresponding to an alphabet $A$ $(|A| =
@@
 \label{probtemplate}
-\end{problem}
+\end{prob}
 
@@
 
-\begin{problem}[{\sc BasicCellPhoneTyping}]~\\
+\begin{prob}[{\sc BasicCellPhoneTyping}]~\\
 {\sc Instance}: A set of letters corresponding to an alphabet $A$ $(|A| =
@@
 \label{bcpt}
-\end{problem}
+\end{prob}
 
-Unlike almost every other problem dealing with key assignment, we can construct
-an efficient algorithm for this problem.  We can construct a provable optimal
+\begin{figure}
+\begin{algorithm}
+\Algorithm{GreedyBasicCellPhoneTyping}~$(A, k, W)$\+\\
+    // $A$ is the set of letters \\
+    // $k$ is the number of keys \\
+    // $W$ is a set of pairs of words and their corresponding frequencies \\
+    lettercount $\gets$ new\_map() \\
+    \For $c \in A$\+\\
+        lettercount$[c] \gets 0$\-\\
+    \For (word, frequency) $\in W$\+\\
+        \For $c \in$ word \+\\
+            lettercount$[c] \gets $ lettercount$[c] +$frequency\-\-\\
+    ordered $\gets \left[ ({\rm lettercount}[c], c)~\For c \in {\rm lettercount} \right]$\\
+    sort(ordered)\\
+    keys $\gets$ a array of length $k$ where each element is an empty list\\
+    \For $i \gets 0 \ldots length($ordered$)$\+\\
+        (count, char) $\gets$ ordered$[i]$\\
+        append(char, keys$[i mod k]$)\-\\
+    \Return keys
+\end{algorithm}
+\caption{The greedy algorithm for Problem~\ref{bcpt}.}
+\label{greedyalg}
+\end{figure}
+
+To solve {\sc GreedyBasicCellPhoneTyping} we construct a provably optimal
 greedy algorithm which works in time $O(|W| + |A| \log |A|)$.  Our algorithm is
@@
 all letters which require two button presses, and so on.  If two assignments
-have the same sequence of sets, then one assignment may be transformed into
+have equal spectra, then one assignment may be transformed into
 the other assignment simply by swapping letter between keys without
 increasing or decreasing the total number of button pushes required to enter
-the corpus of words.  If two sequences have equal spectra, then we say their corresponding assignments are isomorphic.
+the corpus of words.   Two assignments are isomorphic iff their spectra are equal.
 
@@
 In an effort to minimize the pain of cellphone keyboard typing, cellular
-telephone manufacturers have created the ``T9'' input method, which attempts to
+telephone manufacturers have created the T9 input method, which attempts to
 guess which letter (of the 3 or 4 possible) is intended when a user hits a
-single key.  Enhancements of this input method also provide ``speculative
-lookahead'' to report, at any given time, what word it is that the user is most
+single key.  Enhancements of this input method also provide \newword{speculative
+lookahead} to report, at any given time, what word it is that the user is most
 likely trying to enter and provide a completion.  Unfortunately, because there
-are fewer keys than there are letters in the English alphabet, there are words
+are fewer keys on the cellphone keyboard than there are letters in the English alphabet, there are words
 which have different spellings but the same input sequence.  As an example, in
-the traditional mapping of keys to characters ($abc\to 2$, $def\to3$, etc),
+the traditional mapping of keys to characters
 both ``me'' and ``of'' have the input sequence ''63''.  Two words with the same
@@
 ``next match'' button, then the user will actually have to type in
-``4663******'' to type in the word ``goof''.
+``4663******'' to type in the word ``goof'', for a total of 10 button presses --- exactly the same number of button presses required using the default layout and the basic input method, and three more button presses than is required when using the basic input method with an optimized layout.
 
 When two words have the same input sequence (neglecting the *'s at the end) then these words are \newword{t9onyms}.
-When two or more words are t9onyms, then the less-popular words require extra keypresses, raising the expected number of keypresses to type in a random selection of words from the english dictionary.
+When two or more words are t9onyms, then the less-popular words require extra keypresses, raising the expected number of keypresses to type in our corpus.
 In order to type a given word, one must press one button for every character in the word, followed by pressing the * key as many times as there are t9onyms which are more likely than the desired wored.
@@
 The number of keystrokes a word requires is equal to the number of letters in the word, plus the number of t9onyms which are more frequent than the word.
-Formally, we define the {\sc MinimalKeystrokes} problem as:
+Formally, we extend Problem~\ref{probtemplate} and define the {\sc MinimumT9Keystrokes} problem as:
 
-\begin{theorem}[{\sc MinimumKeystrokes}]~\\
+\begin{prob}[{\sc MinimumT9Keystrokes}]~\\
 \label{thm:minstrokes}
-{\sc Instance} An alphabet $A$, a set of words and their associated probabilities $W$, and a set of keys $K$ ($|A| > |K|$).\\
+{\sc Instance} An alphabet $A$, a set of words and their associated frequencies $W$, and a number of keys $k$ ($|A| > k$).\\
 ~\\
-{\sc Question} What is the mapping of $A \to K$ which minimizes the expected number of characters to type a word from $W$?
-\end{theorem}
+{\sc Question} What is the partion $P$ of $A$ into $k$ sets which minimizes the total button presses required to enter the entire corpus using the T9 input method?
+\end{prob}
 
@@
 
-\begin{theorem}~\\
-{\sc Instance} An alphabet $A$, a set of words and their associated probabilities $W$, a set of keys $K$ ($|A| > |K|$), and a number $f$.\\
+\begin{prob}~\\
+{\sc Instance} An alphabet $A$, a set of words and their associated frequencies $W$, a number of keys $k$ ($|A| > k$), and a number $t$.\\
 ~\\
-{\sc Question} Is there a mapping of $A \to K$ in which the expected number of characters to type a word from $W$ is less than $f$?
-\end{theorem}
+{\sc Question} Is there a partition of $A$ into $k$ sets in which the total number of keystrokes required to enter the entire corpus using the T9 input method requires no more than $t$ button presses?
+\end{prob}
 
@@
 In order to decide this, we define the problem {\sc MinimumKeystrokePartition} as:
-\begin{theorem}[{\sc MinimumKeystrokePartition}]~\\
+\begin{prob}[{\sc MinimumKeystrokePartition}]~\\
 \label{thm:minpartition}
@@
 {\sc Question} What is the mapping $f$ of $A \to K$ which minimizes the expected number of characters to type a word from $W$, and where $f(a_1) = 2$ and if $f(a_i) = k$, then either $f(a_{i+1}) = k$ or $f(a_{i+1}) = k+1$?
-\end{theorem}
+\end{prob}
 
 As before, the corresponding decision problem is
-\begin{theorem}~\\
+\begin{prob}~\\
 {\sc Instance} A sequence of letters $A = \{a_1, a_2, \ldots \}$, a set of words and their associated probabilities $W$, a set of keys $K = \{2, 3, 4, \ldots \}$ ($|A| > |K|$), and a parameter $e$.\\
@@
 {\sc Question} Is there a mapping $f$ of $A \to K$ in which the expected number of characters to type a word from $W$ is less than $e$, and where $f(a_1) = 2$ and if $f(a_i) = k$, then $f(a_{i+1}) \in \{k, k+1\}$?
-\end{theorem}
+\end{prob}
 

### 2010-01-16 article.tex
Wording changes

@@
 {\sc Instance}: A set of letters corresponding to an alphabet $A$ $(|A| =
-n)$, a number of keys $k$, an input method $IN$, and a set of
+n)$, a number of keys $k$, an input method $\mathrm{IN}$, and a set of
 tuples of words and frequencies $W$.  The frequencies in $W$ are integers,
@@
 will minimize
-$$\sum_{(w,f)\in W} f*IN(w,P)$$
+$$\sum_{(w,f)\in W} f*\mathrm{IN}(w,P)$$
 \label{probtemplate}
@@
 Over the course of this paper we consider three different real-world
-schemes for $IN$ (basic typing, T9, and predictive T9), and for each
+schemes for $\mathrm{IN}$ (basic typing, T9, and predictive T9), and for each
 variant that is proven \NP-hard, we consider the restriction on $P$ that
@@
 3.  Equivalently, if we denote the number of button presses required to type
-letter $a$ with partition $P$ as $IN_P(a)$, then we want to find the $P$ that
-minimizes $$\sum_{(w,f) \in W}\sum_{c\in w} IN_P(c) * f$$
+letter $a$ with partition $P$ as $\mathrm{IN}_P(a)$, then we want to find the $P$ that
+minimizes $$\sum_{(w,f) \in W}\sum_{c\in w} \mathrm{IN}_P(c) * f$$
 \label{bcpt}

### 2010-01-16 article.tex
Typo fix
Fixed an egregious write-o in accounting for the number of sequences

@@
 
-If we examine a cell phone keyboard (Figure~\ref{keypic}) then we can see that
+If we examine a cell phone keyboard (Fig.~\ref{keypic}) then we can see that
 there are some terrible choices being made.  The frequently-occurring letters
@@
 \end{algorithm}
-\caption{The greedy algorithm for Problem~\ref{bcpt}.}
+\caption{The greedy algorithm for Prob.~\ref{bcpt}.}
 \label{greedyalg}
@@
 
-To solve {\sc GreedyBasicCellPhoneTyping} we construct a provably optimal
+To solve {\sc BasicCellPhoneTyping} we construct a provably optimal
 greedy algorithm which works in time $O(|W| + |A| \log |A|)$.  Our algorithm is
-detailed in Figure~\ref{greedyalg}, and involves creating a histogram of the
+detailed in Fig.~\ref{greedyalg}, and involves creating a histogram of the
 letters, and then assigning letters to keys round-robin style in the order from
@@
 The number of keystrokes a word requires is equal to the number of letters in the word, plus the number of t9onyms which are more frequent than the word.
-Formally, we extend Problem~\ref{probtemplate} and define the {\sc MinimumT9Keystrokes} problem as:
+Formally, we extend Prob.~\ref{probtemplate} and define the {\sc MinimumT9Keystrokes} problem as:
 
@@
 Because our problems are \NP-hard, we move on to combinatorial enumeration and experimental results.
-The number of partitions of a set of size $n$, which we denote $p(n)$ is actually relatively small.
-In particular, $p(26) = 2,436$\footnote{\url{http://www.wolframalpha.com/input/?i=PartitionsP[26]}}, which is easily enumerated and tested.
+The number of partitions of a sequence of size $n$ into $k$ subsequences is equal to $\binom{n-1}{k-1}$, and in the particular case of the 26 letter alphabet and the eight keys available on a cell phone keyboard, we find that there are $\binom{26-1}{8-1} = 480,700$ possible sequences.
 

### 2010-01-20 article.tex
Trying for a new aticle.tex with \Question and \Instance defined

@@
 \newcommand{\NP}{\ensuremath{\mathcal{NP}}}
+\newcommand{\Instance}{\Instance}
+\newcommand{\Question}{~\\
+\Question}
 \spnewtheorem{prob}[theorem]{Problem}{\bfseries}{\itshape}
@@
 \begin{prob}[{\sc MinimumKeystrokes}]~\\
-{\sc Instance}: A set of letters corresponding to an alphabet $A$ $(|A| =
+\Instance\ A set of letters corresponding to an alphabet $A$ $(|A| =
 n)$, a number of keys $k$, an input method $\mathrm{IN}$, and a set of
@@
 treat $IN$ as a function which, given a partition of $A$ and a word
-$w$, returns how many keystrokes are required to type $w$.~\\
-        ~\\
-{\sc Question}: What is the best partition of $A$ into $k$ sets, such that the
+$w$, returns how many keystrokes are required to type $w$.
+
+\Question\ What is the best partition of $A$ into $k$ sets, such that the
 total number of keystrokes to type every word in $W$ its associated frequency
@@
 \begin{prob}[{\sc BasicCellPhoneTyping}]~\\
-{\sc Instance}: A set of letters corresponding to an alphabet $A$ $(|A| =
+\Instance\ A set of letters corresponding to an alphabet $A$ $(|A| =
 n)$, a number of keys $k$, and a set of
@@
         ~\\
-{\sc Question}: What is the best partition $P$ of $A$ into $k$ sequences, such
+\Question\ What is the best partition $P$ of $A$ into $k$ sequences, such
 that the total number of button presses to type every word in $W$ its
@@
 \label{thm:minstrokes}
-{\sc Instance} An alphabet $A$, a set of words and their associated frequencies $W$, and a number of keys $k$ ($|A| > k$).\\
+\Instance\ An alphabet $A$, a set of words and their associated frequencies $W$, and a number of keys $k$ ($|A| > k$).\\
 ~\\
-{\sc Question} What is the partion $P$ of $A$ into $k$ sets which minimizes the total button presses required to enter the entire corpus using the T9 input method?
+\Question\ What is the partion $P$ of $A$ into $k$ sets which minimizes the total button presses required to enter the entire corpus using the T9 input method?
 \end{prob}
@@
 \begin{prob}~\\
-{\sc Instance} An alphabet $A$, a set of words and their associated frequencies $W$, a number of keys $k$ ($|A| > k$), and a number $t$.\\
+\Instance\ An alphabet $A$, a set of words and their associated frequencies $W$, a number of keys $k$ ($|A| > k$), and a number $t$.\\
 ~\\
-{\sc Question} Is there a partition of $A$ into $k$ sets in which the total number of keystrokes required to enter the entire corpus using the T9 input method requires no more than $t$ button presses?
+\Question\ Is there a partition of $A$ into $k$ sets in which the total number of keystrokes required to enter the entire corpus using the T9 input method requires no more than $t$ button presses?
 \end{prob}
@@
 less than $f$ in time proportional to the total length of all the words in $W$.
+In order to prove completeness, however, we first prove the \NP-completeness of an intermediate problem: {\sc UniquePathColoring}.
 
-It is highly likely this problem is \NP-complete, due to a reduction from one of the optimal trie problems.  But I'm not there yet.
+\begin{prob}[{\sc UniquePathColoring}]~\\
+\label{upcolor}
+\Instance\ A graph $G=(V,E)$, a set of paths $P$, and a parameter $k$.\\
+~\\
+\Question\
+\end{prob}
 
@@
 \label{thm:minpartition}
-{\sc Instance} A sequence of letters $A = \{a_1, a_2, \ldots \}$, a set of words and their associated probabilities $W$, and a set of keys $K = \{2, 3, 4, \ldots \}$ ($|A| > |K|$).\\
-~\\
-{\sc Question} What is the mapping $f$ of $A \to K$ which minimizes the expected number of characters to type a word from $W$, and where $f(a_1) = 2$ and if $f(a_i) = k$, then either $f(a_{i+1}) = k$ or $f(a_{i+1}) = k+1$?
+\Instance\ A sequence of letters $A = \{a_1, a_2, \ldots \}$, a set of words and their associated probabilities $W$, and a set of keys $K = \{2, 3, 4, \ldots \}$ ($|A| > |K|$).
+
+\Question\ What is the mapping $f$ of $A \to K$ which minimizes the expected number of characters to type a word from $W$, and where $f(a_1) = 2$ and if $f(a_i) = k$, then either $f(a_{i+1}) = k$ or $f(a_{i+1}) = k+1$?
 \end{prob}
@@
 \begin{prob}~\\
-{\sc Instance} A sequence of letters $A = \{a_1, a_2, \ldots \}$, a set of words and their associated probabilities $W$, a set of keys $K = \{2, 3, 4, \ldots \}$ ($|A| > |K|$), and a parameter $e$.\\
-~\\
-{\sc Question} Is there a mapping $f$ of $A \to K$ in which the expected number of characters to type a word from $W$ is less than $e$, and where $f(a_1) = 2$ and if $f(a_i) = k$, then $f(a_{i+1}) \in \{k, k+1\}$?
+\Instance\ A sequence of letters $A = \{a_1, a_2, \ldots \}$, a set of words and their associated probabilities $W$, a set of keys $K = \{2, 3, 4, \ldots \}$ ($|A| > |K|$), and a parameter $e$.
+
+\Question\ Is there a mapping $f$ of $A \to K$ in which the expected number of characters to type a word from $W$ is less than $e$, and where $f(a_1) = 2$ and if $f(a_i) = k$, then $f(a_{i+1}) \in \{k, k+1\}$?
 \end{prob}

### 2010-01-20 article.tex
Woot!  Proof going in!

@@
 \newcommand{\NP}{\ensuremath{\mathcal{NP}}}
-\newcommand{\Instance}{\Instance}
+\newcommand{\Instance}{{\sc Instance~}}
 \newcommand{\Question}{~\\
-\Question}
+{\sc Question~}}
 \spnewtheorem{prob}[theorem]{Problem}{\bfseries}{\itshape}
@@
 tuples of words and frequencies $W$.  The frequencies in $W$ are integers,
-and the words are made up of solely of elements of $A$. \\
-        ~\\
+and the words are made up of solely of elements of $A$.
+
 \Question\ What is the best partition $P$ of $A$ into $k$ sequences, such
@@
 described layout is the only legacy preserving layout in this paper, and
-therefore should be considered the most practical suggestion, despite the fact
-that its algorithm is among the least FUN.  It also has the added benefit of
-not angering any of the organizations behind the numbers 1-800-FLOWERS,
-1-800-THRIFTY, and 1-800-MARINES}.  After setting up this baseline for the easy
-problem, we turn our attention to optimizing cellphone keyboards for predictive
-input methods.
+therefore should be considered the most practical suggestion.  It also has the
+added benefit of not angering any of the organizations behind the numbers
+1-800-FLOWERS, 1-800-THRIFTY, and 1-800-MARINES}.  After setting up this
+baseline for the easy problem, we turn our attention to optimizing cellphone
+keyboards for predictive input methods.
 
@@
 guess which letter (of the 3 or 4 possible) is intended when a user hits a
-single key.  Enhancements of this input method also provide \newword{speculative
-lookahead} to report, at any given time, what word it is that the user is most
-likely trying to enter and provide a completion.  Unfortunately, because there
-are fewer keys on the cellphone keyboard than there are letters in the English alphabet, there are words
-which have different spellings but the same input sequence.  As an example, in
-the traditional mapping of keys to characters
-both ``me'' and ``of'' have the input sequence ''63''.  Two words with the same
-input sequence force the user to press a third button to cycle through the
-possibilities in order from most likely (``of'') to least likely (``me'').
-Even worse: ``home'', ``good'', ``gone'', ``hood'', ``hoof'', ``hone'', and
-``goof'' all have the input sequence ``4663''.  If the * key is used as the
-``next match'' button, then the user will actually have to type in
-``4663******'' to type in the word ``goof'', for a total of 10 button presses --- exactly the same number of button presses required using the default layout and the basic input method, and three more button presses than is required when using the basic input method with an optimized layout.
+single key.  Enhancements of this input method also provide
+\newword{speculative lookahead} to report, at any given time, what word it is
+that the user is most likely trying to enter and provide a completion.
+Unfortunately, because there are fewer keys on the cellphone keyboard than
+there are letters in the English alphabet, there are words which have different
+spellings but the same input sequence.  As an example, in the traditional
+mapping of keys to characters both ``me'' and ``of'' have the input sequence
+``63''.  Two words with the same input sequence force the user to press a third
+button to cycle through the possibilities in order from most likely (``of'') to
+least likely (``me'').  Even worse: ``home'', ``good'', ``gone'', ``hood'',
+``hoof'', ``hone'', and ``goof'' all have the input sequence ``4663''.  If
+the * key is used as the ``next match'' button, then the user will actually
+have to type in ``4663******'' to type in the word ``goof'', for a total of 10
+button presses --- exactly the same number of button presses required using the
+default layout and the basic input method, and three more button presses than
+is required when using the basic input method with an optimized layout.
 
-When two words have the same input sequence (neglecting the *'s at the end) then these words are \newword{t9onyms}.
-When two or more words are t9onyms, then the less-popular words require extra keypresses, raising the expected number of keypresses to type in our corpus.
-In order to type a given word, one must press one button for every character in the word, followed by pressing the * key as many times as there are t9onyms which are more likely than the desired wored.
-Because typing on a cellphone keyboard is already an unpleasant experience, we would like to minimize the expected number of keystrokes.
-The number of keystrokes a word requires is equal to the number of letters in the word, plus the number of t9onyms which are more frequent than the word.
-Formally, we extend Prob.~\ref{probtemplate} and define the {\sc MinimumT9Keystrokes} problem as:
+When two words have the same input sequence (neglecting the *'s at the end)
+then these words are \newword{t9onyms}.  When two or more words are t9onyms,
+then the less-popular words require extra keypresses, raising the expected
+number of keypresses to type in our corpus.  In order to type a given word, one
+must press one button for every character in the word, followed by pressing
+the * key as many times as there are t9onyms which are more likely than the
+desired wored.  Because typing on a cellphone keyboard is already an unpleasant
+experience, we would like to minimize the expected number of keystrokes.  The
+number of keystrokes a word requires is equal to the number of letters in the
+word, plus the number of t9onyms which are more frequent than the word.
+Formally, we extend Prob.~\ref{probtemplate} and define the {\sc
+MinimumT9Keystrokes} problem as:
 
@@
 \label{thm:minstrokes}
-\Instance\ An alphabet $A$, a set of words and their associated frequencies $W$, and a number of keys $k$ ($|A| > k$).\\
-~\\
+\Instance\ An alphabet $A$, a set of words and their associated frequencies $W$, and a number of keys $k$ ($|A| > k$).
+
 \Question\ What is the partion $P$ of $A$ into $k$ sets which minimizes the total button presses required to enter the entire corpus using the T9 input method?
@@
 \begin{prob}~\\
-\Instance\ An alphabet $A$, a set of words and their associated frequencies $W$, a number of keys $k$ ($|A| > k$), and a number $t$.\\
-~\\
+\Instance\ An alphabet $A$, a set of words and their associated frequencies $W$, a number of keys $k$ ($|A| > k$), and a number $t$.
+
 \Question\ Is there a partition of $A$ into $k$ sets in which the total number of keystrokes required to enter the entire corpus using the T9 input method requires no more than $t$ button presses?
@@
 
-Before proceeding, we note that this we note that this problem is in \NP, as
-any assignment may be verified to have and expected number of characters of
-less than $f$ in time proportional to the total length of all the words in $W$.
-In order to prove completeness, however, we first prove the \NP-completeness of an intermediate problem: {\sc UniquePathColoring}.
+Note that this problem is in \NP, as any assignment may be verified to require
+fewer than $f$ button pushes in time proportional to the
+total length of all the words in $W$.  In order to prove completeness, however,
+we first prove the \NP-completeness of an intermediate problem: {\sc
+UniquePathColoring}.
 
@@
 \label{upcolor}
-\Instance\ A graph $G=(V,E)$, a set of paths $P$, and a parameter $k$.\\
-~\\
-\Question\
+\Instance\ A graph $G=(V,E)$, a set of paths $P$, and a parameter $k$.  A path $p$ is a sequence of vertices in which adjacent vertices in $p$ are also adjacent in the edge set $E$.
+
+\Question\ Is there a $k$-coloring of $G$ such that every path $p\in P$ has a unique coloring?
+If we consider the coloring function $\chi(v)$ to map vertices to colors, then we can extend this notation by having $\chi(p)$ map a path $p = [ v_1, v_2, \ldots ]$ to the sequence $\chi(p) = [ \chi(v_1), \chi(v_2), \ldots ]$.
+Is there a coloring $\chi$ of $V$ such that
+    $$|\left\{ \chi(p)~\forall p \in P\right\}| = |P|\enspace ?$$
 \end{prob}
+
+\begin{proof}[{\sc UniquePathColoring} is \NP-complete]
+To prove that {\sc UniquePathColoring} is \NP-complete, we begin by noting that any coloring of $V$ may be verified, in polynomial time, to map each path to a unique sequence of colors.
+Therefore, the problem is in \NP.
+To prove completeness, we reduce from {\sc GraphColoring}\cite{gandj}.
+
+An instance of {\sc GraphColoring } consists of a graph $G=(V,E)$ and a parameter $k$.
+We then ask the question of whether there is a $k$-coloring $\chi$ of the vertices of the graph such that $\forall (u,v)\in E$, $\chi(u) \neq \chi(v)$.
+We transform an instance of {\sc GraphColoring} into an instance of {\sc UniquePathColoring} in the following manner:
+
+Given $G=(V,E)$ and $k$ from {\sc GraphColoring}, we create 
+$$G'=\left(V \cup \{0,1\}, E \cup \{(0,1)\} \cup \{ (v, 0)~ \forall v \in V\} \right)$$
+We then uniquely number each edge in $E$ with the numbers $1 \ldots |E|$.  For each edge $e = (u,v)$ numbered $i$, we create the path set 
+\begin{eqnarray*}
+p_i &=& \{[v,0,b_1(i),b_2(i), b_3(i), \ldots b_{\lceil \log_2 i\rceil}], \\
+    & &  ~[u, 0,b_1(i),b_2(i), b_3(i), \ldots b_{\lceil \log_2 i\rceil}]\} \\
+\end{eqnarray*}
+where $b_1(i)$ is the first digit of $i$ in binary, $b_2(i)$ is the second digit of $i$ in binary, and so on.  Now we create our path set 
+$$P = \{ [0,1], [1,0] \} \cup \bigcup_{i = 1}^{|E|} p_i \enspace .$$
+We then ask the question of whether $G'$ and $P$ can be unique-path colored using only $k$ colors.
+
+If there is a $k$-coloring $\chi$ of $G$, then we use that coloring to generate
+a $k$-unique-path coloring of $G'$ and $P$ in the following manner: First,
+assign vertices $0$ and $1$ different colors from the range of $\chi$.  Next,
+assign each vertex to the color it received in the $k$-coloring of $G'$.  Now,
+every element of our path set corresponds to a unique color sequence.  $[0,1]$
+and $[1,0]$ are the only sequences of length two, and because we assigned these
+two vertices different colors, $\chi([0,1]) \neq \chi([1,0])$.  If path $x$ and
+path $y$ are not from the same $p_i$, then they have a different sequence of
+zeroes and ones.  Therefore, because $0$ and $1$ are assigned different colors,
+the only possible path that a given path in some $p_i$ might be confused with
+is the other path in that $p_i$.  But each $p_i$ corresponds to a an edge in
+$E$, and we know for all edges in $(u,v) \in E$ that $\chi(u) \neq \chi(v)$.
+Therefore, both paths within a given $p_i$ also have a distinct color sequence.  Therefore, given a $k$-coloring of $G$, we can generate a $k$-unique-path coloring of $G'$ and $P$.
+
+Now we prove that given a $k$-unique-path coloring of $G'$ and $P$ we can create a $k$-coloring of $G$.
+
+\end{proof}
 

### 2010-01-21 article.tex
Small changes.  Looking good!

@@
 
-\begin{figure}
+\begin{figure}[t]
+\begin{center}
 \includegraphics[width=2in]{phonekeys.jpg}
+\end{center}
 \caption{A cell phone keyboard}
@@
 
-There are two cases for layer $T_i$.  In the first case, $T_i < k$, and we may
-remove any element from a later layer and create a strictly better layout.  In
+There are two cases for layer $T_i$.  In the first case, $|T_i| < k$, and we may
+remove any element from a later layer place it in layer $T_i$ and create a strictly better layout.  In
 the second case, there is some letter $a$ such that $a \in T_i$ and $a \not\in
@@
 $T$ was not an optimal layout, and we have reached our contradiction.
+\qed
 \end{proof}
@@
 the optimal layout in this new scheme is to place the letters of a given key in
-sorted order.  Thus, the keyboard layout changes to $2\to[acb], 3\to[edf],
-4\to[ihg], 5\to[lkj], 6\to[onm], 7\to[srpq], 8\to[tuv], 9\to[wyxz]$ and
-requires 678,547,463 button presses to enter the whole corpus.  This represents
-a $28.43\%$ savings, but has the added benefit that it does not change which
-key is mapped to which set of letters.  This makes it \newword{legacy
-preserving}, as creating a keyboard with this layout will not invalidate such
-advertising gems as 1-800-FLOWERS.  Thus, we can speed up users by $28.43\%$
-(neglecting inter-letter pauses) and not undermine 50 years of advertising.
-This layout represents perhaps the most plausible layout yet\footnote{The
-described layout is the only legacy preserving layout in this paper, and
-therefore should be considered the most practical suggestion.  It also has the
-added benefit of not angering any of the organizations behind the numbers
-1-800-FLOWERS, 1-800-THRIFTY, and 1-800-MARINES}.  After setting up this
-baseline for the easy problem, we turn our attention to optimizing cellphone
-keyboards for predictive input methods.
+sorted order, according to frequency.  Thus, the keyboard layout changes to
+$2\to[acb], 3\to[edf], 4\to[ihg], 5\to[lkj], 6\to[onm], 7\to[srpq], 8\to[tuv],
+9\to[wyxz]$ and requires 678,547,463 button presses to enter the whole corpus.
+This represents a $28.43\%$ savings, but has the added benefit that it does not
+change which key is mapped to which set of letters.  This makes it
+\newword{legacy preserving}, as creating a keyboard with this layout will not
+invalidate such advertising gems as 1-800-FLOWERS.  Thus, we can speed up users
+by $28.43\%$ (neglecting inter-letter pauses) and not undermine 50 years of
+advertising.  This layout represents perhaps the most plausible layout
+yet\footnote{The described layout is the only legacy preserving layout in this
+paper, and therefore should be considered the most practical suggestion.  It
+also has the added benefit of not angering any of the organizations behind the
+numbers 1-800-FLOWERS, 1-800-THRIFTY, and 1-800-MARINES}.  After setting up
+this baseline for the easy problem, we turn our attention to optimizing
+cellphone keyboards for predictive input methods.
 
@@
 Now we prove that given a $k$-unique-path coloring of $G'$ and $P$ we can create a $k$-coloring of $G$.
-
+\qed
 \end{proof}

### 2010-01-22 article.tex
Proven?

@@
 Before we get any farther, let us sketch the basic problem that we will keep
-revising and revisiting throughout this paper.
-\begin{prob}[{\sc MinimumKeystrokes}]~\\
-\Instance\ A set of letters corresponding to an alphabet $A$ $(|A| =
-n)$, a number of keys $k$, an input method $\mathrm{IN}$, and a set of
-tuples of words and frequencies $W$.  The frequencies in $W$ are integers,
-and the words are made up of solely of elements of $A$.  We will
-treat $IN$ as a function which, given a partition of $A$ and a word
-$w$, returns how many keystrokes are required to type $w$.
+revising and revisiting throughout this paper.  \begin{prob}[{\sc
+MinimumKeystrokes}]~\\ \Instance\ A set of letters corresponding to an alphabet
+$A$ $(|A| = n)$, a number of keys $k$, an input method $\mathrm{IN}$, and a set
+of tuples of words and frequencies $W$.  The frequencies in $W$ are integers,
+and the words are made up of solely of elements of $A$.  We will treat $IN$ as
+a function which, given a partition of $A$ and a word $w$, returns how many
+keystrokes are required to type $w$.
 
@@
 
-Over the course of this paper we consider three different real-world
-schemes for $\mathrm{IN}$ (basic typing, T9, and predictive T9), and for each
-variant that is proven \NP-hard, we consider the restriction on $P$ that
-requires that we keep the alphabet in alphabetical order.  In all cases
-where it is computationally feasible we provide results for the case of
-the English language, on 8-key keyboards, using the British National
-Corpus\cite{bnc} (BNC).  One feature of note is that no matter what scheme we
-use, the problem is trivial if the number of keys is not smaller than the
-number of letters in the alphabet ($k \ge |A|$).  Using tiny keyboards is only
-(computationally) difficult when the keyboards don't have enough keys.
+Over the course of this paper we consider three different real-world schemes
+for $\mathrm{IN}$ (basic typing, T9, and predictive T9), and for each variant
+that is proven \NP-hard, we consider the restriction on $P$ that requires that
+we keep the alphabet in alphabetical order.  In all cases where it is
+computationally feasible we provide results for the case of the English
+language, on 8-key keyboards, using the British National Corpus\cite{bnc}
+(BNC).  One feature of note is that no matter what scheme we use, the problem
+is trivial if the number of keys is not smaller than the number of letters in
+the alphabet ($k \ge |A|$).  Using tiny keyboards is only (computationally)
+difficult when the keyboards don't have enough keys.
 
@@
 
-The baseline we compare against is the most painful method of text entry.  I
-imagine that no ``digital native'' actually uses this input method, and that it
-is available on all phones solely that their parents might use it.  In this
+The baseline we compare against is the most painful method of text entry.  In this
 method, to type an `a', the user of the cellphone types the `2' key; to type a
 `b' they type `22', to type `c' they type `222', to type a `d' they type `3',
-to type an `e' they type `33', and so on.  To type a word with multiple
-letters that use the same key, such as `accept', the user must pause
-between keystrokes.  Thus, the full key entry sequence for `accept', using
-`.' to indicate a pause, is `2.22.223378'.
+to type an `e' they type `33', and so on.  To type a word with multiple letters
+that use the same key, such as `accept', the user must pause between
+keystrokes.  Thus, the full key entry sequence for `accept', using `.' to
+indicate a pause, is `2.22.223378'.
 
@@
 \begin{proof}
-We will assume, for simplicity of proof, that all letters occur a different number of times in the corpus.  
-We begin by comparing two assignments of letters to keys by, for each assignment,
-constructing a sequence of sets, or spectrum, $S$.  The set $S_1$ (layer 1) is the set of all
-letters which require a single button press (that is, they are the first
-letter in their sequence on their assigned key), $S_2$ (layer 2) is the set of
-all letters which require two button presses, and so on.  If two assignments
-have equal spectra, then one assignment may be transformed into
-the other assignment simply by swapping letter between keys without
-increasing or decreasing the total number of button pushes required to enter
-the corpus of words.   Two assignments are isomorphic iff their spectra are equal.
+We will assume, for simplicity of proof, that all letters occur a different
+number of times in the corpus.  We begin by comparing two assignments of
+letters to keys by, for each assignment, constructing a sequence of sets, or
+spectrum, $S$.  The set $S_1$ (layer 1) is the set of all letters which require
+a single button press (that is, they are the first letter in their sequence on
+their assigned key), $S_2$ (layer 2) is the set of all letters which require
+two button presses, and so on.  If two assignments have equal spectra, then one
+assignment may be transformed into the other assignment simply by swapping
+letter between keys without increasing or decreasing the total number of button
+pushes required to enter the corpus of words.   Two assignments are isomorphic
+iff their spectra are equal.
 
-Now assume for the sake of contradiction that we have assignment $D$, with spectrum $T$, that is not isomorphic to the greedy solution $G$ with spectrum $S$.
-Note that in spectrum resulting from the greedy algorithm, all letters at layer
-$i$ occur strictly more frequently than the letters at layer $j$, $j > i$.
-Also, in $S$, for all layers $i$ except for the last layer, $|S_i| = k$.
-Because $T$ and $S$ are not isomorphic, in $T$ there must be at least one layer $i$ such that $T_i \neq S_i$.  Let $T_i$ be the first layer for which that is true.
+Now assume for the sake of contradiction that we have assignment $D$, with
+spectrum $T$, that is not isomorphic to the greedy solution $G$ with spectrum
+$S$.  Note that in spectrum resulting from the greedy algorithm, all letters at
+layer $i$ occur strictly more frequently than the letters at layer $j$, $j >
+i$.  Also, in $S$, for all layers $i$ except for the last layer, $|S_i| = k$.
+Because $T$ and $S$ are not isomorphic, in $T$ there must be at least one layer
+$i$ such that $T_i \neq S_i$.  Let $T_i$ be the first layer for which that is
+true.
 
-There are two cases for layer $T_i$.  In the first case, $|T_i| < k$, and we may
-remove any element from a later layer place it in layer $T_i$ and create a strictly better layout.  In
-the second case, there is some letter $a$ such that $a \in T_i$ and $a \not\in
-S_i$.  We choose the least-frequent such $a$.  Because the greedy layout
-algorithm creates layers in order of frequency, we know that the frequency of
-$a$ is less than that of some letter $b$ in layer $T_j$, $j>i$.  Therefore, by
-creating a new layout with $a$ and $b$ swapped between $T_i$ and $T_j$, we have
-strictly decreased the number of button presses.  Therefore, in both cases,
-$T$ was not an optimal layout, and we have reached our contradiction.
-\qed
-\end{proof}
+There are two cases for layer $T_i$.  In the first case, $|T_i| < k$, and we
+may remove any element from a later layer place it in layer $T_i$ and create a
+strictly better layout.  In the second case, there is some letter $a$ such that
+$a \in T_i$ and $a \not\in S_i$.  We choose the least-frequent such $a$.
+Because the greedy layout algorithm creates layers in order of frequency, we
+know that the frequency of $a$ is less than that of some letter $b$ in layer
+$T_j$, $j>i$.  Therefore, by creating a new layout with $a$ and $b$ swapped
+between $T_i$ and $T_j$, we have strictly decreased the number of button
+presses.  Therefore, in both cases, $T$ was not an optimal layout, and we have
+reached our contradiction.  \qed \end{proof}
 
-For the BNC, the optimal layout has the spectrum 
-$$[\{etaoinsr\}, \{hldcumfp\}, \{gwybvkxj\}, \{qz\}]$$
-and any layout with that spectrum requires 638,276,114 button presses to
-entire the entire BNC instead of our original requirements of
-948,049,046.  This represents a savings of $32.67\%$ over our original
-layout, but only for the users who type using this most basic of
-input methods.  Unfortunately, this is the group of users who are likely
-the least adaptable to change, as almost all ``digital natives'' use 
-predictive methods to input text messages\footnote{I have no statistics
-on this except for an informal survey of one of my classes.  In that
-class, every student who used SMS either had a cellphone with a 26+
-key alphabetic keyboard or used a predictive method.}.  In order to place the
-least burden of ``newness'' on these users whilst still decreasing the number
-of button presses, we now consider schemes where the only alteration of the
-keyboard is the rearrangement of the sequence of letters for a given key.
+For the BNC, the optimal layout has the spectrum $$[\{etaoinsr\}, \{hldcumfp\},
+\{gwybvkxj\}, \{qz\}]$$ and any layout with that spectrum requires 638,276,114
+button presses to entire the entire BNC instead of our original requirements of
+948,049,046.  This represents a savings of $32.67\%$ over our original layout,
+but only for the users who type using this most basic of input methods.
+Unfortunately, this is the group of users who are likely the least adaptable to
+change, as almost all ``digital natives'' use predictive methods to input text
+messages\footnote{I have no statistics on this except for an informal survey of
+one of my classes.  In that class, every student who used SMS either had a
+cellphone with a 26+ key alphabetic keyboard or used a predictive method.}.  In
+order to place the least burden of ``newness'' on these users whilst still
+decreasing the number of button presses, we now consider schemes where the only
+alteration of the keyboard is the rearrangement of the sequence of letters for
+a given key.
 
@@
 spellings but the same input sequence.  As an example, in the traditional
-mapping of keys to characters both ``me'' and ``of'' have the input sequence
-``63''.  Two words with the same input sequence force the user to press a third
-button to cycle through the possibilities in order from most likely (``of'') to
-least likely (``me'').  Even worse: ``home'', ``good'', ``gone'', ``hood'',
-``hoof'', ``hone'', and ``goof'' all have the input sequence ``4663''.  If
+mapping of keys to characters both `me' and `of' have the input sequence
+`63'.  Two words with the same input sequence force the user to press a third
+button to cycle through the possibilities in order from most likely (`of') to
+least likely (`me').  Even worse: `home', `good', `gone', `hood',
+`hoof', `hone', and `goof' all have the input sequence `4663'.  If
 the * key is used as the ``next match'' button, then the user will actually
-have to type in ``4663******'' to type in the word ``goof'', for a total of 10
+have to type in `4663******' to type in the word `goof', for a total of 10
 button presses --- exactly the same number of button presses required using the
@@
 \label{thm:minstrokes}
-\Instance\ An alphabet $A$, a set of words and their associated frequencies $W$, and a number of keys $k$ ($|A| > k$).
+\Instance\ An alphabet $A$, a set of words and their associated frequencies
+$W$, and a number of keys $k$ ($|A| > k$).
 
-\Question\ What is the partion $P$ of $A$ into $k$ sets which minimizes the total button presses required to enter the entire corpus using the T9 input method?
+\Question\ What is the partition $P$ of $A$ into $k$ sets which minimizes the
+total button presses required to enter the entire corpus using the T9 input
+method?
 \end{prob}
@@
 \begin{prob}~\\
-\Instance\ An alphabet $A$, a set of words and their associated frequencies $W$, a number of keys $k$ ($|A| > k$), and a number $t$.
+\label{minstrokesdecision}
+\Instance\ A set $A$, a set $W$ of sequences of elements of $A$, a mapping $f$ from sequences in $W$ to the integers, and a number $t$.
 
-\Question\ Is there a partition of $A$ into $k$ sets in which the total number of keystrokes required to enter the entire corpus using the T9 input method requires no more than $t$ button presses?
+\Question\ Is there a partition $P$ of $A$ into $k$ sets such that, if we denote $P(w)$ as the sequence of partitions $[p_i | w_i \in p_i, p_i \in P]$,
+$$\sum_{w\in W} (\mathrm{len}(w)+\mathrm{order}(w,P,W,f)) * f(w) \le t$$
+where len$(w)$ is the length of the sequence $w$ and order$(w,P,W,f)$ is the size of $\left\{s\in W | P(s) = P(w) \land f(s) > f(w)\right\}$, which is the set of sequences which map to the same sequence of partitions, but are strictly more popular.
 \end{prob}
@@
 Note that this problem is in \NP, as any assignment may be verified to require
-fewer than $f$ button pushes in time proportional to the
-total length of all the words in $W$.  In order to prove completeness, however,
-we first prove the \NP-completeness of an intermediate problem: {\sc
-UniquePathColoring}.
+fewer than $f$ button pushes in time proportional to the total length of all
+the words in $W$.  In order to prove completeness, however, we first prove the
+\NP-completeness of an intermediate problem: {\sc UniquePathColoring}.
 
@@
 \label{upcolor}
-\Instance\ A graph $G=(V,E)$, a set of paths $P$, and a parameter $k$.  A path $p$ is a sequence of vertices in which adjacent vertices in $p$ are also adjacent in the edge set $E$.
+\Instance\ A graph $G=(V,E)$, a set of paths $P$, and a parameter $k$.  A path
+$p$ is a sequence of vertices in which adjacent vertices in $p$ are also
+adjacent in the edge set $E$.
 
-\Question\ Is there a $k$-coloring of $G$ such that every path $p\in P$ has a unique coloring?
-If we consider the coloring function $\chi(v)$ to map vertices to colors, then we can extend this notation by having $\chi(p)$ map a path $p = [ v_1, v_2, \ldots ]$ to the sequence $\chi(p) = [ \chi(v_1), \chi(v_2), \ldots ]$.
-Is there a coloring $\chi$ of $V$ such that
+\Question\ Is there a $k$-coloring of $G$ such that every path $p\in P$ has a
+unique coloring?  If we consider the coloring function $\chi(v)$ to map
+vertices to colors, then we can extend this notation by having $\chi(p)$ map a
+path $p = [ v_1, v_2, \ldots ]$ to the sequence $\chi(p) = [ \chi(v_1),
+\chi(v_2), \ldots ]$.  Is there a coloring $\chi$ of $V$ such that 
     $$|\left\{ \chi(p)~\forall p \in P\right\}| = |P|\enspace ?$$
@@
 
-\begin{proof}[{\sc UniquePathColoring} is \NP-complete]
-To prove that {\sc UniquePathColoring} is \NP-complete, we begin by noting that any coloring of $V$ may be verified, in polynomial time, to map each path to a unique sequence of colors.
-Therefore, the problem is in \NP.
-To prove completeness, we reduce from {\sc GraphColoring}\cite{gandj}.
+\begin{theorem}{\sc UniquePathColoring} is \NP-complete.\end{theorem}
+\begin{proof}
+To prove that {\sc UniquePathColoring} is \NP-complete, we begin by noting that
+any coloring of $V$ may be verified, in polynomial time, to map each path to a
+unique sequence of colors.  Therefore, the problem is in \NP.  To prove
+completeness, we reduce from {\sc GraphColoring}\cite{gandj}.
 
-An instance of {\sc GraphColoring } consists of a graph $G=(V,E)$ and a parameter $k$.
-We then ask the question of whether there is a $k$-coloring $\chi$ of the vertices of the graph such that $\forall (u,v)\in E$, $\chi(u) \neq \chi(v)$.
-We transform an instance of {\sc GraphColoring} into an instance of {\sc UniquePathColoring} in the following manner:
+An instance of {\sc GraphColoring } consists of a graph $G=(V,E)$ and a
+parameter $k$.  We then ask the question of whether there is a $k$-coloring
+$\chi$ of the vertices of the graph such that $\forall (u,v)\in E$, $\chi(u)
+\neq \chi(v)$.  We transform an instance of {\sc GraphColoring} into an
+instance of {\sc UniquePathColoring} in the following manner:
 
-Given $G=(V,E)$ and $k$ from {\sc GraphColoring}, we create 
-$$G'=\left(V \cup \{0,1\}, E \cup \{(0,1)\} \cup \{ (v, 0)~ \forall v \in V\} \right)$$
-We then uniquely number each edge in $E$ with the numbers $1 \ldots |E|$.  For each edge $e = (u,v)$ numbered $i$, we create the path set 
+Given $G=(V,E)$ and $k$ from {\sc GraphColoring}, we create $$G'=\left(V \cup
+\{0,1\}, E \cup \{(0,1)\} \cup \{ (v, 0)~ \forall v \in V\} \right)$$ We then
+uniquely number each edge in $E$ with the numbers $1 \ldots |E|$.  For each
+edge $e = (u,v)$ numbered $i$, we create the path set 
 \begin{eqnarray*}
@@
 \end{eqnarray*}
-where $b_1(i)$ is the first digit of $i$ in binary, $b_2(i)$ is the second digit of $i$ in binary, and so on.  Now we create our path set 
-$$P = \{ [0,1], [1,0] \} \cup \bigcup_{i = 1}^{|E|} p_i \enspace .$$
-We then ask the question of whether $G'$ and $P$ can be unique-path colored using only $k$ colors.
+where $b_1(i)$ is the first digit of $i$ in binary, $b_2(i)$ is the second
+digit of $i$ in binary, and so on.  Now we create our path set $$P = \{ [0,1],
+[1,0] \} \cup \bigcup_{i = 1}^{|E|} p_i \enspace .$$ We then ask the question
+of whether $G'$ and $P$ can be unique-path colored using only $k$ colors.
 
@@
 $E$, and we know for all edges in $(u,v) \in E$ that $\chi(u) \neq \chi(v)$.
-Therefore, both paths within a given $p_i$ also have a distinct color sequence.  Therefore, given a $k$-coloring of $G$, we can generate a $k$-unique-path coloring of $G'$ and $P$.
+Therefore, both paths within a given $p_i$ also have a distinct color sequence.
+Therefore, given a $k$-coloring of $G$, we can generate a $k$-unique-path
+coloring of $G'$ and $P$.
 
-Now we prove that given a $k$-unique-path coloring of $G'$ and $P$ we can create a $k$-coloring of $G$.
+Now we prove that given a $k$-unique-path coloring of $G'$ and $P$ we can
+create a $k$-coloring of $G$.  Given a $k$-unique path coloring, we are
+guaranteed that no path in $P$ has the exact same coloring as any other path in
+$P$.  This means that, for all $i$, the two paths in $p_i$ are distinct.  The
+only difference between the two paths in $p_i$ is in the first vertex of the
+path.  Therefore, for every $p_i$, the vertices of corresponding edge in must
+be given distinct colors, and this is true for all $i$ and $e \in E$.
+Therefore, for all $(u,v)\in E$, the color assigned to $u$ is distinct from the
+color assigned to $v$.  Therefore, from a $k$-unique-path coloring of $G'$ and
+$P$, we have a $k$-coloring of $G=(V,E)$.
+
 \qed 
 \end{proof}
+
+Now that we have proven {\sc UniquePathColoring} to be \NP-complete, we
+immediately use it in a reduction.  
+\begin{theorem}{\sc MinimumT9Keystrokes} (the decision problem form,
+Problem~\ref{minstrokesdecision}) is \NP-complete.  \end{theorem}
 

### 2010-01-24 genetic.cpp
Genetic algorithm for finding good T9 layouts.

@@ -1,0 +1,165 @@
+#include <algorithm>
+#include <iostream>
+#include <fstream>
+#include <map>
+#include <string>
+#include <vector>
+#include <sys/time.h>
+
+#define POPULATION 32
+#define GENERATIONS 20
+
+using namespace std;
+
+typedef pair<int, string> pis;
+vector<pis> words;
+long long total;
+
+bool piscompare(pis a, pis b) { return a.first > b.first; }
+
+long long quality(vector<string> &partition)
+{
+    map<char,string> rmap;
+    int i = 0;
+    string buttons = "23456789";
+    for (vector<string>::iterator s = partition.begin();
+            s != partition.end();
+            ++s, ++i) {
+        for (string::iterator c = s->begin(); c != s->end(); ++c) {
+            rmap[*c] = buttons[i];
+        }
+    }
+
+    map<string, vector<pis> > typing;
+    for (vector<pis>::iterator fw = words.begin();
+            fw != words.end();
+            ++fw) {
+        int freq = fw->first;
+        string word = fw->second;
+        string presses = "";
+        for (string::iterator c = word.begin(); c != word.end(); ++c) {
+            presses += rmap[*c];
+        }
+
+        if (typing.count(presses) == 0)
+            typing[presses] = vector<pis>();
+        typing[presses].push_back(*fw);
+    }
+
+    for (map<string, vector<pis> >::iterator w = typing.begin();
+            w != typing.end();
+            ++w) {
+        sort(w->second.begin(), w->second.end(), piscompare);
+    }
+
+    long long totalpresses = 0;
+    for (map<string, vector<pis> >::iterator w = typing.begin();
+            w != typing.end();
+            ++w) {
+        int wordsize = w->first.size();
+        for (int i = 0; i < w->second.size(); i++) {
+            totalpresses += (wordsize + i) * w->second[i].first;
+        }
+    }
+    return totalpresses;
+}
+
+void printpart(vector<string> &p)
+{
+    for (vector<string>::iterator i = p.begin();
+            i != p.end();
+            ++i) 
+        cout << *i << " ";
+    cout << endl;
+}
+
+vector<string> randompartition(string in)
+{
+    vector<string> p;
+    for (int i = 0; i < 8; i++)
+        p.push_back("");
+
+    while (in.size() > 0) {
+        int pos = random() % in.size();
+        int key = random() % 8;
[…]

### 2010-01-24 article.tex
Submission numero uno and maybe numero final.  A reread tomorrow, but I'm uploading this one now.

@@
 $\{2\to\{abc\}, 3\to\{def\}\ldots\}$.  Our overall goal is to minimize the
-number of buttons that must be pressed to enter an aferage message in English.
-We prove that the problem in almost all of its variations is \NP\ hard, but
-describe several mappings which improve the standard one.  With luck, one of
-these new mappings will achieve success similar to that of the Dvorak layout
-for computer keyboards.
+expected number of buttons that must be pressed to enter a message in English.
+Some variations of the problem are efficiently solvable, either by being small
+instances or by being in \ensuremath{P}, but the most realistic version of the
+problem is \NP\ hard.  In order to prove \NP-completeness, we describe a new
+graph coloring problem {\sc UniquePathColoring}.  We also provide numerical
+results for the English language on a standard corpus which describe several
+mappings that improve upon the standard one.  With luck, one of these new
+mappings will achieve success similar to that of the Dvorak layout for computer
+keyboards.
 \end{abstract}
@@
 a baseline of how many keystrokes are required to enter the entirety of our
-corpus of words $W$, we count the total number of occurences of each letter,
+corpus of words $W$, we count the total number of occurrences of each letter,
 and multiply that number of occurrences by the number of button pushes
@@
         (count, char) $\gets$ ordered$[i]$\\
-        append(char, keys$[i mod k]$)\-\\
+        append(char, keys$[i\mod k]$)\-\\
     \Return keys
@@
 cellphone with a 26+ key alphabetic keyboard or used a predictive method.}.  In
-order to place the least burden of ``newness'' on these users whilst still
+order to place the least burden of ``newness'' on these users while still
 decreasing the number of button presses, we now consider schemes where the only
@@
 single key.  Enhancements of this input method also provide
-\newword{speculative lookahead} to report, at any given time, what word it is
+speculative lookahead to report, at any given time, what word it is
 that the user is most likely trying to enter and provide a completion.
@@
 then these words are \newword{t9onyms}.  When two or more words are t9onyms,
-then the less-popular words require extra keypresses, raising the expected
-number of keypresses to type in our corpus.  In order to type a given word, one
+then the less-popular words require extra key presses, raising the expected
+number of key presses to type in our corpus.  In order to type a given word, one
 must press one button for every character in the word, followed by pressing
 the * key as many times as there are t9onyms which are more likely than the
-desired wored.  Because typing on a cellphone keyboard is already an unpleasant
+desired word.  Because typing on a cellphone keyboard is already an unpleasant
 experience, we would like to minimize the expected number of keystrokes.  The
@@
 $$\sum_{w\in W} (\mathrm{len}(w)+\mathrm{order}(w,P,W,f)) * f(w) \le t$$
-where len$(w)$ is the length of the sequence $w$ and order$(w,P,W,f)$ is the size of $\left\{s\in W | P(s) = P(w) \land f(s) > f(w)\right\}$, which is the set of sequences which map to the same sequence of partitions, but are strictly more popular.
+where len$(w)$ is the length of the sequence $w$ and order$(w,P,W,f)$ is the size of $\left\{s\in W | P(s) = P(w) \land f(s) \le f(w) \land s < w \right\}$, which is the set of sequences which map to the same sequence of partitions, but are not more popular, and are lexicographically smaller than $w$.
 \end{prob}
@@
 Note that this problem is in \NP, as any assignment may be verified to require
-fewer than $f$ button pushes in time proportional to the total length of all
+fewer than $t$ button pushes in time proportional to the total length of all
 the words in $W$.  In order to prove completeness, however, we first prove the
@@
 
+Note that this proof implies not just that {\sc UniquePathColoring} is \NP-complete, but that it is \NP-complete eve when we restrict ourself to just 3 colors, because the number of colors in the {\sc UniquePathColoring} is exactly equal to the number of colors in the instance of {\sc GraphColoring}, and 3-coloring a graph is an \NP-complete problem\cite{3color}.
+
 Now that we have proven {\sc UniquePathColoring} to be \NP-complete, we
 immediately use it in a reduction.  
-\begin{theorem}{\sc MinimumT9Keystrokes} (the decision problem form,
+\begin{theorem}{\sc MinimumT9Keystrokes} (as specified in
 Problem~\ref{minstrokesdecision}) is \NP-complete.  \end{theorem}
 
-However, our arbitrary remappings of the keyboard may be quickly deemed unusable.
-Therefore, we define a refinement of the problem, in which the characters are required to be kept in alphabetical order as a sequence, and our only choices are about how to divide the subsequence into key assignments.
-The default layout can be described as the partition $abc|def|ghi|jkl|mno|pqrs|tuv|wxyz$.
-Is this partition optimal?
-In order to decide this, we define the problem {\sc MinimumKeystrokePartition} as:
+\begin{proof}
+As we noted on page~\pageref{minstrokesdecision}, {\sc MinimumT9Keystrokes} is
+in \NP.  To prove completeness, we reduce from {\sc UniquePathColoring}.  An
+instance of {\sc UniquePathColoring} consists of a graph $G=(V,E)$, a set of
+paths $P$, and a parameter $k$.  We then ask the question of whether there is a
+vertex coloring of $G$ using $k$ colors such that each path in $P$ ends up with
+a unique sequence of colors.
+
+We transform an instance of {\sc UniquePathColoring} into an instance of {\sc
+MinimumT9Keystrokes} in the following way:  We make $A$ be the set of vertices
+$V$, and we make $W$ be the set of paths $P$, and we assign each of these new
+`words' a frequency of 1.  We then ask the question of whether there exists a
+partition of $A$ into $k$ sets such that the total number of button presses
+required is no greater than $t=\sum_{w\in W} \mathrm{len}(w)$.
+
+Now we prove that if there is a {\sc UniquePathColoring} using $k$ colors, then
+there exists a solution to the corresponding instance of {\sc
+MinimumT9Keystrokes}.  In particular, a $k$-unique-path coloring if $G$ give us
+a partition of the vertices $V$ into $k$ sets.  We map each set to a single key
+in our solution to {\sc MinimumT9Keystrokes}.  Because each path has a unique
+coloring, each word in the corresponding {\sc MinimumT9Keystrokes} has a unique
+pattern of button presses.  Therefore, the number of more popular words with
+the same keystrokes as a given word is exactly zero.  Thus, because each word
+in the corpus has a frequency of one, to type in the entire corpus only
+requires exactly as many keystrokes as there are words in the corpus, and
+therefore, the $k$ unique path coloring of $V$ corresponds exactly to an
+assignment of letters to keys such that the total number of keystrokes is
+exactly $\sum_{w\in W} \mathrm{len}(w)$, which is exactly $t$.
+
+Next, we prove that a solution to the {\sc MinimumT9Keystrokes} problem
+instance implies a solution to the corresponding {\sc UniquePathColoring}
+problem.  If there exists a partition of $A$ into $k$ partitions such that
+$$\sum_{w\in W} (\mathrm{len}(w)+\mathrm{order}(w,P,W,f))*f(w) \le t =
+\sum_{w\in W} \mathrm{len}(w)$$ then, because $f(w) = 1$, $ \forall w \in W$,
+it must be true that $\mathrm{order}(w,P,W,f) = 0$,  $\forall w \in W$.
+Therefore, every word $w \in W$ has a unique sequence of button presses if we
+partition $A$ into $k$ partitions according to $P$.  To generate our
+$k$-unique-path coloring of $G$, we color each set in $P$ a single unique
+color.  Because each word in $w$ has a unique sequence of partitions, each path
+has a unique coloring, and therefore we can color the corresponding instance of
+$G$ using $k$ colors.
+Thus, a partition of $V$ into $k$ partitions which allows the corpus to be entered in less than $t$ keystrokes implies a $k$-unique path coloring of $G$.
+Therefore, because {\sc UniquePathColoring} can be reduced to {\sc MinimumT9Keystrokes} and is in \NP, {\sc MinimumT9Keystrokes} is \NP-complete.
+\qed
+\end{proof}
+
+Again we note that this proof of \NP-completeness is of a relatively strong form: just like {\sc UniquePathColoring}, our problem remains \NP-complete even if we restrict ourselves to $k=3$ (only 3 buttons on the cell phone keyboard).
+
+Because our problem is \NP-complete, we will not try to find a general solution.
+In our specific instance, with 26 letters and an eight key keyboard (1 and 0 are reserved for other uses),  we must choose the best layout from among $$\binom{26}{8} \binom{18}{8} \binom{10}{8} = 3,076,291,325,250$$ different possibilities.
+Given that, currently, it takes around a tenth of a second to count the number of button presses required to enter the BNC, an exhaustive search will take 307,629,132,525 seconds, or 9.75 millennia.
+Therefore, we have two remaining avenues of attack: find the best answer possible using stochastic methods, or solving a simpler problem.
+Using stochastic methods (a simple genetic algorithm), the best layout we found was
+$$\{\{eb\}, \{lcd\}, \{sh\}, \{zmx\}, \{fayg\}, \{toj\}, \{unpv\}, \{irkwq\}\}$$
+which required 441,612,049 button presses to enter the entire corpus.
+As compared to the requirements of the default layout (443,374,079), this represents a savings of less than 1\%, and is thus not a very attractive alternative to existing layouts.
+This low amount of improvement, while not good news for our proposed reordering, is excellent news for all current cell phone users: cell phone keyboard layout seem relatively efficient for T9 text entry!
+
+\subsection{Alphabetic Order}
+
+Our arbitrary remappings of the keyboard may be quickly deemed
+unusable, simply because it is almost impossible to tell where a given key is located without memorizing the entire new layout.
+Therefore, we define a refinement of the keyboard layout problem, in which the
+characters are required to be kept in alphabetical order as a sequence, and our
+only choices are about how to divide the subsequence into key assignments.  The
+default layout can be described as the partition
+$abc|def|ghi|jkl|mno|pqrs|tuv|wxyz$.  Is this partition optimal?  In order to
+decide this, we define the problem {\sc MinimumKeystrokePartition} as:
 \begin{prob}[{\sc MinimumKeystrokePartition}]~\\
@@
 
-As before, the corresponding decision problem is
-\begin{prob}~\\
-\Instance\ A sequence of letters $A = \{a_1, a_2, \ldots \}$, a set of words and their associated probabilities $W$, a set of keys $K = \{2, 3, 4, \ldots \}$ ($|A| > |K|$), and a parameter $e$.
+The number of partitions of a sequence of size $n$ into $k$ subsequences is equal to $\binom{n-1}{k-1}$, and in the particular case of the 26 letter alphabet and the eight keys available on a cell phone keyboard, we find that there are $\binom{26-1}{8-1} = 480,700$ possible sequences.
+This means that, in the case of our corpus which requires tenths of a second to test against a proposed solution, we can test every possible sequence in just 480,700 seconds.
+Therefore, our final result comes from simply generating and testing every single partition of the alphabet into 8 sets.
+We found that the best partition was
+$$\{\{ab\}, \{cde\}, \{fghij\}, \{klm\}, \{nop\}, \{qrs\}, \{t\}, \{uvwxyz\}\}$$
+which required 442,717,436 button presses, yet again for a savings of less than 1\% over the default layout.
 
-\Question\ Is there a mapping $f$ of $A \to K$ in which the expected number of characters to type a word from $W$ is less than $e$, and where $f(a_1) = 2$ and if $f(a_i) = k$, then $f(a_{i+1}) \in \{k, k+1\}$?
-\end{prob}
+\section{Conclusion}
 
-Again, we note that this decision problem is in \NP, as any proposed partition may be verified, in polynomial time, to have an expected value less than $e$.
+In this paper, we have developed three problems based on the idea of making cell phone keyboard more efficient for typing text messages.
+Different text entry methods yield problems with very different levels of solvability.
+If we restrict ourselves to the basic text entry method, then a greedy algorithm will work for finding the optimal layout.
+As a matter of fact, the greedy algorithm works whether we want to restrict ourselves to rearranging the order of letters on a single key, or to rearranging all letters on all keys.
+On the other hand, if we try to optimize against the more complex T9 input method, then we find that the problem is \NP-complete.
+After proving completeness, we gave numerical results which indicate that it will be difficult (or perhaps impossible) to significantly improve on the default layout of letters.
+This stands in sharp contrast to the simpler input method, which we were able to improve by more than 27\% just by reordering the letters on the keys, but never moving a letter from one key to another.
 
-\section{Exhaustive Search}
+We have also discovered a new \NP-complete problem {\sc UniquePathColoring}, which may be of use in other contexts.
+In particular, it is rare in the literature to find an \NP-complete problem that contains both partitions and sequences, which is why we had to invent a new one here.
+The hope, as always, is that this problem will prove useful to others.
 
-Because our problems are \NP-hard, we move on to combinatorial enumeration and experimental results.
-The number of partitions of a sequence of size $n$ into $k$ subsequences is equal to $\binom{n-1}{k-1}$, and in the particular case of the 26 letter alphabet and the eight keys available on a cell phone keyboard, we find that there are $\binom{26-1}{8-1} = 480,700$ possible sequences.
+We left as future work any exploration of the issue of needing to pause between
+letters in the basic typing method, and while there do exist even more
+sophisticated input methods, we have left them unmentioned and unexamined.
+Optimizing a keyboard for typing in the basic text method was relatively straightforward, but, in the T9 input method, optimally using a cell phone keyboard is \NP hard.
 

### 2010-03-15 article.tex
Style fixes.  (thanks style-check.rb !)

@@
 \begin{abstract}
-Sending text messages on cell phones which only contain the keys 0-9 and \# and
-* can be a painful experience.  We consider the problem of designing an optimal
-mapping of numbers to sets of letters to act as an alternative to the standard
-$\{2\to\{abc\}, 3\to\{def\}\ldots\}$.  Our overall goal is to minimize the
-expected number of buttons that must be pressed to enter a message in English.
-Some variations of the problem are efficiently solvable, either by being small
-instances or by being in \ensuremath{P}, but the most realistic version of the
-problem is \NP\ hard.  In order to prove \NP-completeness, we describe a new
-graph coloring problem {\sc UniquePathColoring}.  We also provide numerical
+Sending text messages on cell phones which only contain the keys 0 through 9
+and \# and * can be a painful experience.  We consider the problem of designing
+an optimal mapping of numbers to sets of letters to act as an alternative to
+the standard $\{2\to\{abc\}, 3\to\{def\}\ldots\}$.  Our overall goal is to
+minimize the expected number of buttons that must be pressed to enter a message
+in English.  Some variations of the problem are efficiently solvable, either by
+being small instances or by being in \ensuremath{P}, but the most realistic
+version of the problem is \NP\ hard.  To prove \NP-completeness, we describe a
+new graph coloring problem {\sc UniquePathColoring}.  We also provide numerical
 results for the English language on a standard corpus which describe several
@@
 alphabet can be a painful task.  There are a plethora of input schemes which
-attempt to make this task easier, but the one thing they all have in common is
+attempt to make this task easier, but the one thing they have in common is
 that all of these input methods use the standard mapping of numeric keys to
@@
          8\to\{tuv\}, 9\to\{wxyz\}\}$.
-In this paper, we consider schemes of rearranging the numbers on the keys to
+We will consider schemes of rearranging the numbers on the keys to
 make messages easier to type.  Most variants of this problem turn out to be
@@
 Before we get any farther, let us sketch the basic problem that we will keep
-revising and revisiting throughout this paper.  \begin{prob}[{\sc
+revising and revisiting.  \begin{prob}[{\sc
 MinimumKeystrokes}]~\\ \Instance\ A set of letters corresponding to an alphabet
@@
 of tuples of words and frequencies $W$.  The frequencies in $W$ are integers,
-and the words are made up of solely of elements of $A$.  We will treat $IN$ as
+and the words are made up of solely of elements of $A$.  We will treat $\mathrm{IN}$ as
 a function which, given a partition of $A$ and a word $w$, returns how many
@@
 
-Over the course of this paper we consider three different real-world schemes
+Here, we consider three different real-world schemes
 for $\mathrm{IN}$ (basic typing, T9, and predictive T9), and for each variant
@@
 the alphabet ($k \ge |A|$).  Using tiny keyboards is only (computationally)
-difficult when the keyboards don't have enough keys.
+difficult when the keyboards do not have enough keys.
 
@@
 
-If we examine a cell phone keyboard (Fig.~\ref{keypic}) then we can see that
-there are some terrible choices being made.  The frequently-occurring letters
+If we examine a cell phone keyboard (Fig.~\ref{keypic}) then we observe some terrible design choices.  The frequently-occurring letters
 'r' and 's' require more keystrokes than 'q'!  Surely, a better designed
@@
 the users spending $8.7\%$ less time entering their text messages.  In an
-effort to find out how low this can go, we define the problem {\sc
+effort to discover how great these savings can be, we define the problem {\sc
 BasicCellPhoneTyping} based on the template problem on
@@
 spectrum $T$, that is not isomorphic to the greedy solution $G$ with spectrum
-$S$.  Note that in spectrum resulting from the greedy algorithm, all letters at
+$S$.  In a spectrum resulting from the greedy algorithm, all letters at
 layer $i$ occur strictly more frequently than the letters at layer $j$, $j >
@@
 one of my classes.  In that class, every student who used SMS either had a
-cellphone with a 26+ key alphabetic keyboard or used a predictive method.}.  In
-order to place the least burden of ``newness'' on these users while still
+cellphone with a 26+ key alphabetic keyboard or used a predictive method.}. 
+To place the least burden of ``newness'' on these users while still
 decreasing the number of button presses, we now consider schemes where the only
@@
 9\to[wyxz]$ and requires 678,547,463 button presses to enter the whole corpus.
-This represents a $28.43\%$ savings, but has the added benefit that it does not
-change which key is mapped to which set of letters.  This makes it
+This layout represents a $28.43\%$ savings, and it has the added benefit that
+it does not change which key is mapped to which set of letters.  This makes it
 \newword{legacy preserving}, as creating a keyboard with this layout will not
@@
 telephone manufacturers have created the T9 input method, which attempts to
-guess which letter (of the 3 or 4 possible) is intended when a user hits a
-single key.  Enhancements of this input method also provide
-speculative lookahead to report, at any given time, what word it is
-that the user is most likely trying to enter and provide a completion.
-Unfortunately, because there are fewer keys on the cellphone keyboard than
-there are letters in the English alphabet, there are words which have different
-spellings but the same input sequence.  As an example, in the traditional
-mapping of keys to characters both `me' and `of' have the input sequence
-`63'.  Two words with the same input sequence force the user to press a third
-button to cycle through the possibilities in order from most likely (`of') to
-least likely (`me').  Even worse: `home', `good', `gone', `hood',
-`hoof', `hone', and `goof' all have the input sequence `4663'.  If
-the * key is used as the ``next match'' button, then the user will actually
-have to type in `4663******' to type in the word `goof', for a total of 10
-button presses --- exactly the same number of button presses required using the
-default layout and the basic input method, and three more button presses than
-is required when using the basic input method with an optimized layout.
+guess which letter (of the three or four possible) is intended when a user hits
+a single key.  Enhancements of this input method also provide speculative
+lookahead to report, at any given time, what word it is that the user is most
+likely trying to enter and provide a completion.  Unfortunately, because there
+are fewer keys on the cellphone keyboard than there are letters in the English
+alphabet, there are words which have different spellings but the same input
+sequence.  As an example, in the traditional mapping of keys to characters both
+`me' and `of' have the input sequence `63'.  Two words with the same input
+sequence force the user to press a third button to cycle through the
+possibilities in order from most likely (`of') to least likely (`me').  Even
+worse: `home', `good', `gone', `hood', `hoof', `hone', and `goof' all have the
+input sequence `4663'.  If the * key is used as the ``next match'' button, then
+the user will have to type in `4663******' to type in the word `goof',
+for a total of 10 button presses --- exactly the same number of button presses
+required using the default layout and the basic input method, and three more
+button presses than is required when using the basic input method with an
+optimized layout.
 
@@
 then the less-popular words require extra key presses, raising the expected
-number of key presses to type in our corpus.  In order to type a given word, one
+number of key presses to type in our corpus.  To type a given word, one
 must press one button for every character in the word, followed by pressing
@@
 
-Note that this problem is in \NP, as any assignment may be verified to require
+This problem is in \NP, as any assignment may be verified to require
 fewer than $t$ button pushes in time proportional to the total length of all
-the words in $W$.  In order to prove completeness, however, we first prove the
+the words in $W$.  To prove completeness, however, we first prove the
 \NP-completeness of an intermediate problem: {\sc UniquePathColoring}.
@@
 the only possible path that a given path in some $p_i$ might be confused with
-is the other path in that $p_i$.  But each $p_i$ corresponds to a an edge in
+is the other path in that $p_i$.  But each $p_i$ corresponds to an edge in
 $E$, and we know for all edges in $(u,v) \in E$ that $\chi(u) \neq \chi(v)$.
@@
 
-Note that this proof implies not just that {\sc UniquePathColoring} is \NP-complete, but that it is \NP-complete eve when we restrict ourself to just 3 colors, because the number of colors in the {\sc UniquePathColoring} is exactly equal to the number of colors in the instance of {\sc GraphColoring}, and 3-coloring a graph is an \NP-complete problem\cite{3color}.
+This proof implies not just that {\sc UniquePathColoring} is \NP-complete, but that it is \NP-complete even when we restrict ourself to just 3 colors, because the number of colors in the {\sc UniquePathColoring} is exactly equal to the number of colors in the instance of {\sc GraphColoring}, and 3-coloring a graph is an \NP-complete problem\cite{3color}.
 
@@
 paths $P$, and a parameter $k$.  We then ask the question of whether there is a
-vertex coloring of $G$ using $k$ colors such that each path in $P$ ends up with
+vertex coloring of $G$ using $k$ colors where each path in $P$ ends up with
 a unique sequence of colors.
@@
 
-Again we note that this proof of \NP-completeness is of a relatively strong form: just like {\sc UniquePathColoring}, our problem remains \NP-complete even if we restrict ourselves to $k=3$ (only 3 buttons on the cell phone keyboard).
+This proof of \NP-completeness is of a relatively strong form: just like {\sc UniquePathColoring}, our problem remains \NP-complete even if we restrict ourselves to $k=3$ (only 3 buttons on the cell phone keyboard).
 
@@
 In our specific instance, with 26 letters and an eight key keyboard (1 and 0 are reserved for other uses),  we must choose the best layout from among $$\binom{26}{8} \binom{18}{8} \binom{10}{8} = 3,076,291,325,250$$ different possibilities.
-Given that, currently, it takes around a tenth of a second to count the number of button presses required to enter the BNC, an exhaustive search will take 307,629,132,525 seconds, or 9.75 millennia.
+Given that, currently, it takes around a tenth of a second to count the button presses required to enter the BNC, an exhaustive search will take 307,629,132,525 seconds, or 9.75 millennia.
 Therefore, we have two remaining avenues of attack: find the best answer possible using stochastic methods, or solving a simpler problem.
@@
 default layout can be described as the partition
-$abc|def|ghi|jkl|mno|pqrs|tuv|wxyz$.  Is this partition optimal?  In order to
+$abc|def|ghi|jkl|mno|pqrs|tuv|wxyz$.  Is this partition optimal?  To
 decide this, we define the problem {\sc MinimumKeystrokePartition} as:
@@
 
-In this paper, we have developed three problems based on the idea of making cell phone keyboard more efficient for typing text messages.
+We have developed three problems based on the idea of making cell phone keyboard more efficient for typing text messages.
 Different text entry methods yield problems with very different levels of solvability.
 If we restrict ourselves to the basic text entry method, then a greedy algorithm will work for finding the optimal layout.
-As a matter of fact, the greedy algorithm works whether we want to restrict ourselves to rearranging the order of letters on a single key, or to rearranging all letters on all keys.
+Even better, the greedy algorithm works whether we want to restrict ourselves to rearranging the order of letters on a single key, or to rearranging all letters on all keys.
 On the other hand, if we try to optimize against the more complex T9 input method, then we find that the problem is \NP-complete.
 After proving completeness, we gave numerical results which indicate that it will be difficult (or perhaps impossible) to significantly improve on the default layout of letters.
-This stands in sharp contrast to the simpler input method, which we were able to improve by more than 27\% just by reordering the letters on the keys, but never moving a letter from one key to another.
+This stands in sharp contrast to the simpler input method, where we were able to improve by more than 27\% just by reordering the letters on the keys, but never moving a letter from one key to another.
 
@@
 We left as future work any exploration of the issue of needing to pause between
-letters in the basic typing method, and while there do exist even more
-sophisticated input methods, we have left them unmentioned and unexamined.
-Optimizing a keyboard for typing in the basic text method was relatively straightforward, but, in the T9 input method, optimally using a cell phone keyboard is \NP hard.
+letters in the basic typing method, and, while there do exist even more
+sophisticated input methods\cite{hcimethods}, we have left them unmentioned and unexamined.
+Optimizing a keyboard for the basic text input method was relatively straightforward, but, for the T9 input method, optimally using a cell phone keyboard is \NP hard.
 

### 2010-05-31 slides/talk.tex
Small changes.  I like how it reads.  Also Talk v0.1 is in the bag.

\documentclass{beamer}
\usetheme{manhattan}  % Now it's a beamer presentation with the Manhattan College theme!

\usepackage{url}
\usepackage{graphicx}
\usepackage{../salgorithm}
\usepackage{verbatim}
\usepackage{amsthm, amsfonts, amsmath, amssymb, amsxtra}

\newcommand{\NP}{\ensuremath{\mathcal{NP}}}
% Make a new command that will make a new subsection and a frame with the same title
\newcommand{\fst}[2]{\subsection{#1}\frame{\frametitle{#1} #2}}

\title{Using Cell Phone Keyboards is \NP\ Hard }
\author{Peter Boothe}
\date{2 June 2010\\
FUN With Algorithms}
\institute[Manhattan College]{
    \url{peter.boothe@manhattan.edu}\\
    Mathematics and Computer Science \\
    Manhattan College}

\begin{document}
\frame{\titlepage}

\section{Introduction}

\fst{So many letters \ldots}{
\begin{itemize}
    \item People want to use phones to send text messages
    \item English has 26 letters, and each letter cas a lower-case and capital form, and we should not forget numbers, puctuation, and whitespace characters.
    \item Other languages have even more!\footnote{Can you name a language with fewer letters than English?\\}
\end{itemize}
}

\fst{\ldots so few keys}{
\begin{tabular}{c p{2.5in}}
\includegraphics[width=1.5in]{../phonekeys.jpg} & 
\vspace{-2in}
\begin{itemize}
    \item A computer keyboard usually has more than 100 keys.
    \item A phone keyboard has, depending on the model, as few as 8 keys which may be used to input letters.
\end{itemize}
\end{tabular}
}

\fst{Two Input Methods}{
    The old default mapping of numbers to letter sequences:
\begin{center}
\begin{tabular}{l l l}
1 $\to$ [] & 2 $\to$ [a,b,c] & 3 $\to$ [d,e,f] \\
4 $\to$ [g,h,i] & 5 $\to$ [j,k,l] & 6 $\to$ [m,n,o] \\
7 $\to$ [p,q,r,s] & 8 $\to$ [t,u,v] & 9 $\to$ [w,x,y,z]
\end{tabular}
\end{center}

\vspace{.25in}
\begin{description}
    \item[Standard] Press the appropriate key until your letter appears.  e.g. Press 8 two times to get a `u'
    \item[T-9] Load a word-frequency dictionary for the language into the phone, and only require that people type one key for each character.  Display the most popular word with a given keypress sequence.
\end{description}
}

\fst{Our question}{
    Both of these input methods assume the old default mapping.  If we throw that assumption out, then, for each typing method: \\
\vspace{.5in}
\LARGE{\it What is the mapping of numbers to letters that minimizes the number of keypresses required to type a message?}
}

\fst{Our question, formalized}{
\begin{description}
\item[\sc Instance] A number of keys, $k$, an alphabet $A$, an input method $\mathrm{IN}: (word, partition) \to {\Bbb N}$, and a set of words and associated frequencies $W$.  
\item[\sc Question] What is the partition $P$ of $A$ into $k$ sequences which minimizes
$$\sum_{(w,f)\in W} f \times\mathrm{IN}(w,P)$$
?
\end{description}
}

\section{Standard Input Method}
\fst{The Standard Input Method}{
\begin{itemize}
    \item Press a key as many times as needed to made the desired letter appear
    e.g. ``fun'' is 3338866, requiring a total of 7 button presses
    \item Word-independent
    \item Easy to understand
    \item Inefficient  e.g. $7 \to [p,q,r,s]$ means you have to go through `q' to get to `r' or `s'
\end{itemize}
}
\fst{Greedy Algorithm}{
\begin{itemize}
    \item Sort the alphabet by letter frequency.
    \item Assign letters to keys in round-robin order.
    \item Proven optimal via an exchange argument (see paper)
    \item Rearranges the letters on the keys
    \item Possibly a very confusing layout
\end{itemize}
}
\fst{Legacy-Preserving Greedy Algorithm}{
\begin{itemize}
    \item Sort each key's letters by frequency.  Reorder the letters on each key in decreasing frequency order.
    \item Using the British National Corpus, this reduces button presses by 28.43\% 
    \item Less confusing than remapping the whole keyboard
    \item Just fixing the 7 key will save you $\sim$10\%
\end{itemize}
}
\fst{An Open Letter}{
Dear Cell Phone Manufacturers,\\
~\\
\hspace{.25in}Please change your keyboard layout to:\\
\begin{center}
\begin{tabular}{l l l}
1 $\to$ [] & 2 $\to$ [a,c,b] & 3 $\to$ [e,d,f] \\
4 $\to$ [i,g,h] & 5 $\to$ [l,k,j] & 6 $\to$ [o,n,m] \\
7 $\to$ [s,r,p,q] & 8 $\to$ [t,u,v] & 9 $\to$ [w,y,x,z]
\end{tabular}
\end{center}
\hspace{.25in}This new layout will require your customers to type 28\% fewer keystrokes when sending an SMS message, but will not break such famous numbers as 1-800-FLOWERS, 1-800-MARINES, or PA-6-5000.  Or at least fix the 7 key.
~\\
\hspace{3in}Sincerely,\\
\hspace{3in}Peter Boothe
}

\section{T-9 Input Method}
\fst{The T-9 Input Method}{
\begin{itemize}
    \item Per-word, not per-character
    \item One key per letter, display the most popular word with that key sequence e.g. `home' is 4663.
    \item In case of collision, cycle through all words with that key sequence in order of popularity using * e.g. `home', `good', `gone', `hood', `hoof', `hone', and `goof' are all 4663, so `goof' is 4663******
    \item Two words with the same sequence are {\bf t9onyms}
\end{itemize}
\begin{center}
    Can we minimize the expected number of * key presses?\\
or, equivalently,\\
    Can we minimize the expected number of t9onyms?
\end{center}
}

\fst{Minimizing T-9 is \NP-complete}{
\begin{itemize}
    \item We have sequences in the problem (the words in the dictionary)
    \item We have partitions in the problem (the assignment of letters to keys)
    \item Couldn't find a NPC problem with both, so invented one
    \item Proof is via reduction from {\sc UniquePathColoring}
\end{itemize}
}

\fst{{\sc UniquePathColoring}}{
\begin{description}
\item[\sc Instance] A graph $G = (V , E)$, a set of paths $P$, and a parameter $k$.
\item[\sc Question] Is there a $k$-coloring of $V$ such that every path in $P$ has a unique coloring?
\end{description}

Note that this problem meets our criteria of having sequences (the paths of $P$ which correspond eventually to the words in $W$) and partitions (the colors of the graph, which correspond eventually to the key assigned to each letter).  {\sc UniquePathColoring} is equivalent to asking if there is an assignment with no t9onyms.
}

\fst{{\sc UniquePathColoring} is \NP-Complete}{
Reduction from {\sc GraphColoring}.\\
~\\
Take the $G=(V,E)$ and $k$ from {\sc GraphColoring}, and construct $G'$.
\begin{eqnarray*}
G' = & ( & V \cup \{0, 1\}, \\
& & E \cup \{(1,1), (0,1), (0,0)\} \cup \{ (v,0)~\forall v \in V \}\\
& ) &
\end{eqnarray*}
}

\frame{
\frametitle{{\sc UniquePathColoring} is \NP-Complete (2)}

Next, uniquely number the edges in $E$.
Construct $P$ by making paths $p_i$ for each edge $(u,v)$ in $E$.
\begin{eqnarray*}
p_i = & \{ & [ u, 0, \mathrm{b}_1(i), \mathrm{b}_2(i), \ldots \mathrm{b}_{\lceil \log_2 i \rceil}(i) ],\\
      &    & [ v, 0, \mathrm{b}_1(i), \mathrm{b}_2(i), \ldots \mathrm{b}_{\lceil \log_2 i \rceil}(i) ],\\
      &  \} & 
\end{eqnarray*}
where $\mathrm{b}_j(i)$ is the $j$th binary digit of $i$.

$$P = \bigcup_{1 \le i \le |E|} p_i$$
}

\frame{
\frametitle{{\sc UniquePathColoring} is \NP-Complete (3)}
    Does there exist a {\sc UniquePathColoring} of $G'$ with paths $P$ using no
more than $k$ colors?  If so, then the colors assigned to the vertices in $V$
form a $k$-coloring of $G$.\\
~\\
    Does there exist a $k$-coloring of $G$?  If so, we can construct a {\sc UniquePathColoring} of $G'$ and $P$ using $k$ colors --- simply choose two distinct colors for the new vertices 0 and 1.\\
~\\
$\therefore$ {\sc UniquePathColoring} is \NP-complete
}

\fst{{\sc Minimum-T9onyms} is \NP-complete}{
    The decision problem {\sc Minimum-T9onyms}:
\begin{description}
    \item[\sc Instance] Alphabet $A$, set of words and associated frequencies $W$, number of keys $k$, number $t$.
    \item[\sc Question] Is there a partition of $A$ into $k$ sets such that
\[t \ge \sum_{(w,f)\in W} f\times(\mathrm{len}(w) + \mathrm{order}(w,P,W,f))~~?\]
\end{description}

Note that this exactly conforms to our problem specification from the beginning.
}

\frame{
\frametitle{{\sc Minimum-T9onyms} is \NP-complete (2)}
If we require that there be no t9onyms, then this simplifies to 
\[t \ge \sum_{(w,f)\in W} f\times\mathrm{len}(w)\]
or simply a requirement that order$()$ be zero for all words in $W$.\\
}

\frame{
\frametitle{{\sc Minimum-T9onyms} is \NP-complete (3)}
The reduction from {\sc UniquePathColoring}\ldots\\
\vspace{.25in}
\begin{tabular}{p{2in} | p{1.97in}}
{\sc UniquePathColoring} & {\sc Minimum-T9onyms} \\
\\
{\sc Instance}~~Graph $G=(V,E)$, paths $P$, and $k$ colors &
{\sc Instance}~~Alphabet $V$, words $P$ all with frequency 1, number of keys $k$, and $t = \sum_{p\in P} \mathrm{len}(p)$ \\
~\\
{\sc Question}~~Does there exist a $k$-coloring of $V$ such that every path $p\in P$ has a unique coloring? &
{\sc Question}~~Does there exist a $k$-partition of $V$ such that every word $p\in P$ has order$()$ zero? 
\end{tabular}
\qed
}

\section{Numerical Results for T-9}
\fst{Baseline}{Our baseline comparison was the number of keystrokes required to enter the entire British National Corpus using T-9 with the standard layout.

\vspace{.25in}
T-9, with the default layout, requires 443,374,079 keypresses to enter the entire BNC.}

\fst{Enumerating All Alphabetic Partitions}{
If we restrict ourselves to only the partitions which keep the letters in alphabetic order, then we can enumerate all $\frac{25!}{18!7!}$ partitions and test them over the course of a few days on a modern computer.

\vspace{.25in}
We find that the best partition is
\[\{\{ab\}, \{cde\}, \{fghij\}, \{klm\}, \{nop\}, \{qrs\}, \{t\}, \{uvwxyz\}\} \] This partition requires 442,717,436 keypresses, a savings of 656,643 keypresses, or $0.14\%$. 
}
\fst{Genetic Algorithm Results}{
A very simple genetic/stochastic algorithm was run for a few days.  The best solution found was 
\[\{\{be\},\{cdl\},\{hs\},\{mxz\},\{afgy\},\{jot\},\{npuv\},\{ikqrw\}\}\]
This required 441,612,049 keypresses, a savings of 1,762,030 keypresses, or $0.40\%$.
}

\section{Summary}
\fst{Future Work}{
\begin{itemize}
    \item Predictive T-9
    \item Other input methods
    \item When T-9 gets a misspelled word, it is very bad and confusing.  Can we minimize not just t9onyms, but also the number of words that have a cell-phone Levenshtein distance of 1? of 2? of $k$?
    \item We completely ignored the pauses that the standard method requires when typing two letter in a row which involve the same key
    \item Completely solving the particular case of T-9 with the English language and an 8-key keyboard
\end{itemize}
}
\fst{Take-away Messages}{
\begin{itemize}
    \item Cell phone keyboards can provide algorithmic fun
    \item The standard input method can be optimized with the greedy algorithm
    \item The T-9 input method is NP-Complete to optimize
    \item The existing keyboard seems alright for T-9, but can be improved by 28\% for the standard input method without breaking legacy advertising
    \item Cell phone keyboards are not just hard to use, they can also be \NP-hard to use
\end{itemize}
}

\fst{Thanks!}{\begin{center}{\Huge\bf Questions?}\end{center}}

\end{document}

### 2010-05-31 slides/talk.tex
Changed the frac to a binom

@@
 \fst{Enumerating All Alphabetic Partitions}{
-If we restrict ourselves to only the partitions which keep the letters in alphabetic order, then we can enumerate all $\frac{25!}{18!7!}$ partitions and test them over the course of a few days on a modern computer.
+If we restrict ourselves to only the partitions which keep the letters in alphabetic order, then we can enumerate all $\binom{25}{7}$ partitions and test them over the course of a few days on a modern computer.
 

### 2010-06-30 slides/talk.tex
This is the version that got published.

@@
     \item People want to use phones to send text messages
-    \item English has 26 letters, and each letter cas a lower-case and capital form, and we should not forget numbers, puctuation, and whitespace characters.
+    \item English has 26 letters, and each letter has a lower-case and capital form, and we should not forget numbers, puctuation, and whitespace characters.
     \item Other languages have even more!\footnote{Can you name a language with fewer letters than English?\\}
@@
     \item Completely solving the particular case of T-9 with the English language and an 8-key keyboard
+    \item Universal keyboards --- fixed labelings which are good for multiple languages using the standard typing method (thanks Markus Holzer!)
 \end{itemize}

### 2010-06-30 article.tex
Actually fixed the commit

@@
 \usepackage{salgorithm}
+\usepackage{verbatim}
 
@@
 alphabet can be a painful task.  There are a plethora of input schemes which
-attempt to make this task easier, but the one thing they have in common is
+attempt to make this task easier (e.g. \cite{hcimethods,ordering1,godfest} and many more), but the one thing they have in common is
 that all of these input methods use the standard mapping of numeric keys to
 alphabetic numbers of 
-$\{2\to\{abc\},
-         3\to\{def\}, 4\to\{ghi\}, 5\to\{jkl\}, 6\to\{mno\}, 7\to\{pqrs\},
-         8\to\{tuv\}, 9\to\{wxyz\}\}$.
+$\{2\to\{abc\}, 3\to\{def\}, 4\to\{ghi\}, 5\to\{jkl\}, 6\to\{mno\}, 7\to\{pqrs\}, 8\to\{tuv\}, 9\to\{wxyz\}\}$.
 We will consider schemes of rearranging the numbers on the keys to
@@
 
-Here, we consider three different real-world schemes
-for $\mathrm{IN}$ (basic typing, T9, and predictive T9), and for each variant
-that is proven \NP-hard, we consider the restriction on $P$ that requires that
-we keep the alphabet in alphabetical order.  In all cases where it is
-computationally feasible we provide results for the case of the English
-language, on 8-key keyboards, using the British National Corpus\cite{bnc}
-(BNC).  One feature of note is that no matter what scheme we use, the problem
-is trivial if the number of keys is not smaller than the number of letters in
-the alphabet ($k \ge |A|$).  Using tiny keyboards is only (computationally)
-difficult when the keyboards do not have enough keys.
+We will consider two different real-world schemes for $\mathrm{IN}$ (basic
+typing and T9), and for each variant or sub-variant that is proven \NP-hard, we
+consider the restriction on $P$ that requires that we keep the alphabet in
+alphabetical order.  In all cases where it is computationally feasible we
+provide results for the case of the English language, on 8-key keyboards, using
+the British National Corpus\cite{bnc} (BNC).  One feature of note is that no
+matter what scheme we use, the problem is trivial if the number of keys is not
+smaller than the number of letters in the alphabet ($k \ge |A|$).  Using tiny
+keyboards is only (computationally) difficult when the keyboards do not have
+enough keys.
 
@@
 
-The baseline we compare against is the most painful method of text entry.  In this
-method, to type an `a', the user of the cellphone types the `2' key; to type a
-`b' they type `22', to type `c' they type `222', to type a `d' they type `3',
-to type an `e' they type `33', and so on.  To type a word with multiple letters
-that use the same key, such as `accept', the user must pause between
+The baseline we compare against is the most painful method of text entry.  In
+this method, to type an `a', the user of the cellphone types the `2' key; to
+type a `b' they type `22', to type `c' they type `222', to type a `d' they type
+`3', to type an `e' they type `33', and so on.  To type a word with multiple
+letters that use the same key, such as `accept', the user must pause between
 keystrokes.  Thus, the full key entry sequence for `accept', using `.' to
@@
 \end{center}
-\caption{A cell phone keyboard}
+\caption{A cell phone keyboard. Each key is mapped to a letter sequence: $2 \to (abc), 3\to (def), 4\to (ghi), 5\to (jkl), 6 \to (mno), 7 \to (pqrs), 8\to (tuv), 9\to (wxyz) $}
 \label{keypic}
@@
 
-Initially, we will completely neglect the pauses and concern ourselves solely
-with the number of keystrokes required.  This implies that `a' will always
+We will neglect the pauses and concern ourselves solely with the number of
+keystrokes required.  This layout and input method imply that `a' will always
 require one button press, `b' will always require two, and so on.  Thus, to get
@@
 corpus of words $W$, we count the total number of occurrences of each letter,
-and multiply that number of occurrences by the number of button pushes
-required by that letter.  Running this on the British National Corpus
-using the standard cell phone keyboard layout, we find that entering the
-entirety of the 100,106,029 word occurrences in the corpus would require
-948,049,046 button presses.
+and multiply that number of occurrences by the number of button pushes required
+by that letter.  Running this on the British National Corpus using the standard
+cell phone keyboard layout, we find that entering the entirety of the
+100,106,029 word occurrences in the corpus would require 948,049,046 button
+presses.
 
@@
 on the 7 key, from `pqrs' to `srqp', the number of button presses required
-would drop to 865,118,331 --- a savings of more than $8.7\%$!  If we assume
+would drop to 865,118,331 --- a savings of more than $8.7\%$.  If we assume
 that every button press takes an equal amount of time, then this corresponds
@@
 invalidate such advertising gems as 1-800-FLOWERS.  Thus, we can speed up users
-by $28.43\%$ (neglecting inter-letter pauses) and not undermine 50 years of
-advertising.  This layout represents perhaps the most plausible layout
-yet\footnote{The described layout is the only legacy preserving layout in this
-paper, and therefore should be considered the most practical suggestion.  It
-also has the added benefit of not angering any of the organizations behind the
-numbers 1-800-FLOWERS, 1-800-THRIFTY, and 1-800-MARINES}.  After setting up
-this baseline for the easy problem, we turn our attention to optimizing
-cellphone keyboards for predictive input methods.
+by $28.43\%$ (neglecting inter-letter pauses) and not undermine more than half
+a century of advertising.  This layout represents perhaps the most plausible
+layout yet\footnote{The described layout is the only legacy preserving layout
+in this paper, and therefore should be considered the most practical
+suggestion.  It also has the added benefit of not angering any of the
+organizations behind the numbers 1-800-FLOWERS, 1-800-THRIFTY, and
+1-800-MARINES}.  After setting up this baseline for the easy problem, we turn
+our attention to optimizing cellphone keyboards for predictive input methods.
 
@@
 Given $G=(V,E)$ and $k$ from {\sc GraphColoring}, we create $$G'=\left(V \cup
-\{0,1\}, E \cup \{(0,1)\} \cup \{ (v, 0)~ \forall v \in V\} \right)$$ We then
+\{0,1\}, E \cup \{(0,1), (1,0), (0,0), (1,1)\} \cup \{ (v, 0)~ \forall v \in V\} \right)$$ We then
 uniquely number each edge in $E$ with the numbers $1 \ldots |E|$.  For each
@@
 which required 441,612,049 button presses to enter the entire corpus.
+Our genetic algorithm was quite simple: From an initial gene pool of random layouts, we discarded all but the best layouts (the survivors).
+Then, we added to the gene pool random mutations of the survivors to fill out the gene pool back to its original size, and repeated the process of selection and mutation for some time.
 As compared to the requirements of the default layout (443,374,079), this represents a savings of less than 1\%, and is thus not a very attractive alternative to existing layouts.
-This low amount of improvement, while not good news for our proposed reordering, is excellent news for all current cell phone users: cell phone keyboard layout seem relatively efficient for T9 text entry!
+This low amount of improvement, while not good news for our proposed reordering, is excellent news for all current cell phone users: the default cell phone keyboard layout seems relatively efficient for T9 text entry in English!
 
@@
 
-Our arbitrary remappings of the keyboard may be quickly deemed
-unusable, simply because it is almost impossible to tell where a given key is located without memorizing the entire new layout.
-Therefore, we define a refinement of the keyboard layout problem, in which the
-characters are required to be kept in alphabetical order as a sequence, and our
-only choices are about how to divide the subsequence into key assignments.  The
-default layout can be described as the partition
+Our arbitrary remappings of the keyboard may be quickly deemed unusable, simply
+because it is almost impossible to tell where a given key is located without
+memorizing the entire new layout.  Therefore, we define a refinement of the
+keyboard layout problem, in which the characters are required to be kept in
+alphabetical order as a sequence, and our only choices are about how to divide
+the subsequence into key assignments.  The default layout can be described as
+the partition
 $abc|def|ghi|jkl|mno|pqrs|tuv|wxyz$.  Is this partition optimal?  To
@@
 sophisticated input methods\cite{hcimethods}, we have left them unmentioned and unexamined.
-Optimizing a keyboard for the basic text input method was relatively straightforward, but, for the T9 input method, optimally using a cell phone keyboard is \NP hard.
+Optimizing a keyboard for the basic text input method was relatively straightforward, but, for the T9 input method, optimally using a cell phone keyboard is \NP\ hard.
 
@@
 \bibliography{bibliography}
+
+\newpage
+\appendix
+\section{A Genetic Algorithm for Discovering Efficient T9 Layouts}
+{\tiny
+\verbatiminput{genetic.cpp}
+}
+
+\section{Code for Analyzing All Alphabetic-Order T9 Layouts}
+{\tiny
+\verbatiminput{notrcount.cpp}
+}
 \end{document}
